---
title: "Confidentiality and threat model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidentiality and threat model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
library(masque)
```

# What masque protects, and what it does not

`masque` is **not** a privacy-preserving or differential-privacy tool.
It is a **structurally faithful development surrogate**. Read this
vignette before sharing any `masque` output beyond your own machine.

## Threat model

| Actor holds | Wants to learn | What `masque` protects |
|---|---|---|
| Synthetic only | Original raw values | Numeric exact values (collaborate); treatment vocabulary (collaborate); column identities (collaborate, optional) |
| Synthetic + recipe | Original raw values | Numeric exact values. The combination is **as sensitive as the original**. |
| Recipe only | Original raw values | Nothing useful — recipe is meaningless without synthetic |
| Synthetic + external side info (e.g., publicly registered trial) | Identity of treatments / sites | Only the label vocabulary; the design footprint is preserved by design and is recognisable. Side info wins. |

**The recipe is private.** Never share the recipe alongside the
synthetic. The collaborator workflow assumes only the synthetic crosses
the trust boundary.

## What `masque` does

- Preserves enough structure for pipelines to run unchanged.
- Provides two explicit modes (`local` vs `collaborate`) so the
  privacy / fidelity trade-off is never hidden.
- Records every translation in a private recipe that round-trips a
  finished pipeline back onto the original data.
- Audits its own output via `audit_mask()` and flags realistic leakage
  before sharing.

## What `masque` does not do

- It does not provide differential-privacy guarantees.
- It does not make outputs safe for public release.
- It does not anonymise rare strata, small designs, or operational
  metadata (small site x year combinations, contact names, geolocations).
- It does not rewrite arbitrary pipeline source code.

## Mode comparison

| Knob | `mode = "local"` | `mode = "collaborate"` |
|---|---|---|
| Treatment levels | preserved | opaque alias (`trt_001`) |
| Categorical covariate levels | preserved | opaque alias (`<col>_L001`) |
| `ignore` columns | pass-through | dropped |
| Numeric synthesis | empirical-quantile (may emit observed values) | empirical-quantile + within-resolution jitter; integers stochastically rounded |
| NA mask | preserved cell-by-cell | preserved cell-by-cell with explicit warning |
| `audit_mask()` | on demand | automatic at `mask()` time |
| `print(recipe)` | full info | redacted; explicit `reveal_maps()` required |

## A deliberately leaky fixture

To see `audit_mask()` fire, build a fixture with patterns that trip
the heuristics: a PII-pattern column the user keeps (against
`propose_roles()`'s default), and a categorical covariate with a
frequency-1 level.

```{r leaky-fixture}
set.seed(0)
n  <- 60
df <- data.frame(
  rep            = rep(1:3, each = 20),
  variety        = factor(rep(paste0("V", 1:6), 10)),
  contact_email  = factor(rep(c("a@x", "b@y"), 30)),
  rare_treatment = factor(c("only_one", sample(c("alpha","beta","gamma"),
                                                59, replace = TRUE))),
  yield          = rnorm(60, 5, 1),
  stringsAsFactors = FALSE
)

roles <- propose_roles(df)
roles
```

`contact_email` was auto-flagged `pii_suspected = TRUE` and assigned to
`ignore`. We override that — pretending the user explicitly retains the
column — to see what the audit flags.

```{r leaky-roles}
roles$role[roles$col == "yield"]          <- "outcome"
roles$role[roles$col == "contact_email"]  <- "covariate"
roles$role[roles$col == "rare_treatment"] <- "covariate"
roles
```

```{r leaky-mask}
m <- mask(df, roles, mode = "collaborate", seed = 1)
```

`mask()` printed two warnings: one for the high-leakage PII column
retained as a covariate, one for the frequency-1 categorical level.
The audit is stored on `m@audit` (and accessible via `audit_mask(m)`):

```{r leaky-audit}
audit_mask(m)
```

Both `contact_email` (PII retained) and `rare_treatment` (frequency-1
level) are flagged HIGH. The user gets to decide whether to share the
synthetic anyway — `masque` records the risks in `recipe@warnings`,
warns at construction time, but does not block.

## Operational guidance

1. **Default to collaborate mode** when in doubt. Local mode is for
   owner-only development.
2. **Treat the recipe as you treat the original data.** Same security
   class, same access controls.
3. **Run `audit_mask()` before any sharing.** Even when collaborate
   mode auto-runs it, re-print before handing the synthetic over.
4. **Never override `pii_suspected = TRUE` without thinking.** The
   heuristic is conservative; if it triggers, decide explicitly whether
   to keep the column.
5. **Small designs leak.** Cell counts of 1 in any categorical column
   under collaborate mode are HIGH leakage. Aggregate or drop before
   masking.
6. **Exact-design preservation is recognisable.** A publicly registered
   trial design (site x year x rep) is a fingerprint. The synthetic
   does not hide that. If exact-design recognisability is a concern,
   roll up to higher-level groupings before calling `mask()`.

For internals of the recipe object — what it stores, what it doesn't,
how `print()` redacts — see `vignette("recipe_anatomy")`.
