---
title: "Confidentiality and the threat model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidentiality and the threat model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
library(masque)
set.seed(1)
```

## What masque protects, and what it does not

`masque` is **not** a privacy-preserving or differential-privacy tool.
It is a structurally faithful development surrogate. Read this vignette
before sharing any `masque` output beyond your own machine.

The recipe returned by `mask()` is at least as sensitive as the original
data. The whole design assumes that only the *synthetic* crosses the
trust boundary and the recipe stays with the custodian.

| Actor holds | Wants to learn | What `masque` protects |
|---|---|---|
| Synthetic only | Original raw values | Aliased treatment and categorical vocabularies, jittered numerics, dropped ids and free text, and optionally the column names |
| Synthetic + recipe | Original raw values | Nothing -- the combination is as sensitive as the original |
| Recipe only | Original raw values | Nothing useful -- the recipe is meaningless without the synthetic |
| Synthetic + external side information | Identity of treatments / sites | The label vocabulary and the order it was in. A preserved design footprint or a `keep` column is recognisable, level frequencies are unchanged, and side information wins |

What `masque` does: it preserves enough structure for pipelines to run
unchanged, exposes the privacy-versus-fidelity trade-off through two
explicit modes, records every translation in a private recipe that
round-trips, and audits its own output before you share it.

What `masque` does not do: it gives no differential-privacy guarantee, it
does not make output safe for public release, it does not hide rare
strata, small designs, or operational metadata such as small
site-by-year combinations or contact names, and it does not rewrite
pipeline source code.

## Two axes of control: roles and actions

Since version 0.6.0, every column carries a `role` (what it is) and an
`action` (how deeply it is masked). The mode sets the default action per
role. You override any column you like. The four actions are the privacy
dial:

| Action | Effect | Vocabulary visible? |
|---|---|---|
| `keep` | byte-identical pass-through | yes -- real values published |
| `scramble` | re-simulate numerics and row-permute categoricals / dates | yes -- labels stay, assignment moves |
| `alias` | scramble, then replace labels with opaque codes | no |
| `drop` | column omitted from the synthetic | n/a |

The mode chooses sensible defaults so the common case is safe without
per-column work:

| Knob | `mode = "local"` | `mode = "collaborate"` |
|---|---|---|
| Treatment labels | kept | aliased (`trt_001`) |
| Categorical covariate labels | kept | aliased (`<col>_L001`) |
| Date / time columns | row-permuted with class preserved | row-permuted with class preserved |
| Identifiers (`id`) | kept | dropped |
| Free text (`text`) | kept | dropped |
| Numeric synthesis | empirical-quantile (may emit observed values) | empirical-quantile plus within-resolution jitter, with integers stochastically rounded |
| NA mask | preserved cell-by-cell | preserved cell-by-cell |
| `audit_mask()` | on demand | automatic at `mask()` time |
| `print(recipe)` | redacted | redacted, with explicit `reveal_maps()` to inspect |

## Depth controls

Two further controls sharpen the depth where the defaults are not enough.

**Hiding column names.** Names themselves can identify a column
(`roseworthy_yield`). `mask(..., alias_names = TRUE)` replaces every
retained column name with an opaque code, which the recipe inverts on the
round-trip. Pass a character vector to hide only some names.

**Hiding design structure.** Design columns are byte-identical by
default, which preserves the experimental layout exactly -- and a
publicly registered trial layout is a fingerprint. To keep the structure
but hide the site or block *labels*, set a design column's action to
`alias`:

```{r design-alias}
df <- data.frame(
  site = factor(rep(c("Roseworthy", "Minnipa", "Turretfield"), each = 6)),
  rep = rep(1:3, 6),
  yield = rnorm(18)
)
roles <- propose_roles(df, detect = FALSE)
roles <- set_role(roles, "site", role = "design", action = "alias")
s <- synthetic(mask(df, roles, mode = "collaborate", seed = 1))
unique(as.character(s$site))   # real site names gone
table(s$site)                  # ... but the three-site structure intact
```

This knowingly breaks byte-identity, so it is never a default -- you ask
for it explicitly.

**What an alias does and does not hide.** The alias vocabulary is fixed
and sorted (`trt_001`, `trt_002`, ...), but which level receives which
alias is drawn from a random permutation, so for a six-level column there
are 720 equally likely maps. This matters because the vocabularies in
field research are routinely public: a variety roster, an N-rate ladder,
a site list. Under an order-preserving map, sorting a candidate list
would reconstruct the whole assignment from the synthetic alone. It no
longer does, and the same draw governs the join keys a `mask_set()` shares
across tables.

Two limits go with that, and they are yours to manage:

* **The seed is confidential material.** The permutation comes from the
  seeded stream, so `mask(seed = 1)` reproduces the same map, exactly as
  it reproduces the same synthetic values. Publishing the seed beside the
  synthetic re-opens the inversion for anyone who can also reproduce the
  draw. Keep the seed with the recipe, or use `seed = NULL` where
  reproducibility is not required and take a fresh map each call.
* **Frequencies are not hidden.** Aliasing is a bijection on labels: the
  count of each level is unchanged by design, because the design footprint
  is what makes the clone useful. A vocabulary whose level counts are
  known and distinct is still matchable one level at a time. `audit_mask()`
  reports those counts so the exposure is visible rather than implicit.

An alias therefore raises the cost of re-identification; it is not a
cryptographic commitment, and none of it is a differential-privacy
guarantee.

## The conditional clone: preserving the treatment effect

The default numeric synthesis re-simulates every scrambled numeric column
from a single global Gaussian copula. That preserves each column's
marginal distribution and the global covariance, which is enough to
develop most pipelines -- but it severs the relationship between treatment
and outcome. The synthetic outcomes are drawn from the pooled
distribution and the treatment labels are relabelled independently, so a
causal model fitted on the clone sees no association between an arm and
its response. A pipeline whose whole purpose is to estimate a treatment
effect would silently give the wrong answer on such a clone.

The `conditional = TRUE` argument fixes this. It fits and samples the
copula *within each treatment-by-design stratum* rather than pooling, so a
row's synthetic outcome inherits the location of the treatment that row
carries. The treatment-to-outcome map -- the quantity a causal model
reads -- survives the clone, within sampling tolerance. This is the
data-side analogue of preserving a conditional mean embedding rather than
a pooled marginal.

The contrast is easiest to see on a two-arm trial with a known effect:

```{r conditional-fixture}
set.seed(42)
n <- 600
arm <- factor(rep(c("ctrl", "treat"), each = n / 2))
# A real +5-unit effect of the treated arm on yield.
yield <- 10 + 5 * (arm == "treat") + rnorm(n, sd = 2)
trial <- data.frame(genotype = arm, yield = yield)

roles <- propose_roles(trial, detect = FALSE)
roles <- set_role(roles, "genotype", role = "treatment", action = "scramble")
roles <- set_role(roles, "yield", role = "outcome")

true_effect <- coef(lm(yield ~ genotype, trial))[["genotypetreat"]]
true_effect
```

Cloning both ways from the same seed, then re-estimating the effect on
each clone:

```{r conditional-contrast}
marg <- suppressWarnings(
  mask(trial, roles, mode = "local", seed = 1, conditional = FALSE)
)
cond <- suppressWarnings(
  mask(trial, roles, mode = "local", seed = 1, conditional = TRUE)
)

effect_of <- function(m) {
  coef(lm(yield ~ genotype, synthetic(m)))[["genotypetreat"]]
}

data.frame(
  clone = c("marginal (default)", "conditional"),
  estimated_effect = c(effect_of(marg), effect_of(cond)),
  true_effect = true_effect
)
```

The marginal clone collapses the effect toward zero. The conditional
clone recovers it. Both clones still match the pooled marginal of the
outcome:

```{r conditional-marginal-held}
data.frame(
  source = c("original", "marginal clone", "conditional clone"),
  mean_yield = c(
    mean(trial$yield), mean(synthetic(marg)$yield),
    mean(synthetic(cond)$yield)
  ),
  sd_yield = c(
    sd(trial$yield), sd(synthetic(marg)$yield),
    sd(synthetic(cond)$yield)
  )
)
```

The conditioning columns -- the treatment plus any retained design
columns -- are recorded on the recipe, so the choice is auditable:

```{r conditional-recipe}
recipe(cond)@conditional
recipe(cond)@conditioning_cols
recipe(cond)@conditioning_used
recipe(cond)@fallback_frac
```

Read the last two. `conditioning_cols` is what the call asked for;
`conditioning_used` is the stratum the clone actually got, and
`fallback_frac` is the share of rows that did not get a stratum at all.

They differ because a conditional clone needs enough rows per stratum to
fit a stratum-local copula, and the finest stratum available is rarely the
one that has them. Treatment crossed with every retained design column on
a replicated factorial -- six N rates by three varieties by four
replicates -- is seventy-two cells of one row. Rather than pool that
wholesale, `mask()` walks a **coarsening ladder**: it drops design
columns, the finest first, until the cells hold at least five rows.
Treatment columns are never dropped, because the assignment is the thing
the conditional clone exists to preserve. Whatever is still too thin at
the bottom rung is pooled into a global fallback, as before.

Any coarsening, and any residual pooling, raises a classed
`masque_conditional_degraded` warning naming the columns given up and the
pooled share, and both facts are written onto the recipe. Catch the class
if a pipeline should stop when its clone is less conditional than it
asked for:

```r
withCallingHandlers(
  mask(trial, roles, seed = 1, conditional = TRUE),
  masque_conditional_degraded = function(w) stop(conditionMessage(w))
)
```

Conditional cloning composes with both modes and with `mask_set()` (each
table is stratified by its own treatment and design columns). With no
treatment or design column to condition on at all, the path degrades to
the global copula and raises the same classed warning. Reach for it
whenever the development pipeline estimates an effect, not just a
distribution.

## What the copula carries: monotone, not non-monotone, association

Numeric columns that are kept and re-simulated together share one
Gaussian copula, fitted on the normal scores of their ranks. A Gaussian
copula holds a single correlation per pair, so it reproduces a *monotone*
association -- linear, or any order-preserving curve -- but not a
dependence that a correlation cannot express. A non-monotone relationship,
such as a U-shaped dependence of an outcome on a covariate, reads as
near-zero rank correlation and is reproduced as near-independence.

```{r nonmonotone}
set.seed(7)
n <- 2000
x_lin <- rnorm(n)
y_lin <- 2 * x_lin + rnorm(n, sd = 0.5)     # monotone (linear)
x_u <- runif(n, -3, 3)
y_u <- x_u^2 + rnorm(n, sd = 0.5)           # non-monotone (U-shaped)
d <- data.frame(x_lin, y_lin, x_u, y_u)

s <- synthetic(mask(d, propose_roles(d, detect = FALSE), seed = 1))

# R-squared of a quadratic fit captures association of any curvature.
fit_r2 <- function(x, y) summary(lm(y ~ poly(x, 2)))$r.squared
data.frame(
  pair = c("monotone (y = 2x)", "non-monotone (y = x^2)"),
  original = c(fit_r2(x_lin, y_lin), fit_r2(x_u, y_u)),
  clone = c(fit_r2(s$x_lin, s$y_lin), fit_r2(s$x_u, s$y_u))
)
```

The monotone pair keeps its association almost exactly. The non-monotone
pair loses it entirely, so on the clone the covariate carries no
information about the outcome. `conditional = TRUE` does not repair this.
It preserves the outcome's location within each treatment-by-design
stratum, not the curvature of the outcome's dependence on a continuous
covariate. A development pipeline whose modelling step is non-linear -- a
smoothing spline, a generalised additive model, a tree, an interaction
term -- will therefore see on the synthetic only the monotone part of any
relationship present in the original, and a good fit there is not evidence
the step behaves correctly on the real data. Validate such a step by
round-tripping it onto the original through the recipe, not by trusting
its result on the clone.

## The leakage audit

`audit_mask()` inspects the synthetic against the original and grades the
leakage of each column. In collaborate mode it runs automatically and
warns at `mask()` time. Here is a fixture built to trip it: a
PII-suspected column the user retains against the default, and a
categorical covariate with a frequency-one level.

```{r leaky-fixture}
df <- data.frame(
  rep = rep(1:3, each = 20),
  variety = factor(rep(paste0("V", 1:6), 10)),
  contact_email = factor(rep(c("a@x", "b@y"), 30)),
  rare_treatment = factor(c(
    "only_one",
    sample(c("alpha", "beta", "gamma"), 59, replace = TRUE)
  )),
  yield = rnorm(60, 5, 1),
  stringsAsFactors = FALSE
)

roles <- propose_roles(df, mode = "collaborate")
roles[, c("col", "role", "action", "pii_suspected")]
```

`contact_email` was auto-flagged `pii_suspected` and set to `drop`. We
override it -- pretending the custodian insists on keeping it -- and
make the rare column a covariate, then mask:

```{r leaky-mask}
roles <- set_role(roles, "yield", role = "outcome")
roles <- set_role(roles, "contact_email", role = "covariate", action = "keep")
roles <- set_role(roles, "rare_treatment", role = "covariate")
m <- mask(df, roles, mode = "collaborate", seed = 1)
audit_mask(m)
```

`contact_email` (real values kept across the trust boundary) and
`rare_treatment` (a frequency-one level) are flagged. `masque` responds
on two channels. At construction time, `mask()` raises a classed warning
(`masque_high_leakage`) and records the findings in `recipe@warnings`
and on the object's audit -- the guided `masque()` flow never silences
it. At write time, the package-managed writers (`masque()`'s `out` and
`write_set()`) refuse to write while a HIGH finding stands: nothing is
written, and the flagged columns are listed with the remedy (re-role,
alias, or drop, then mask again). A custodian who has reviewed the
findings can pass `allow_high = TRUE` to write anyway. The override is
raised as a `masque_high_override` warning and recorded in the recipe's
warnings, so the exception stays auditable.

Beyond that gate the release decision stays with the custodian --
whether a synthetic table is appropriate for a given collaborator,
environment, or jurisdiction depends on context the package cannot see.
`masque` informs that decision. It does not make it.

## Multi-table sets

When several tables share a key, `mask_set()` aliases that key
*identically across all of them* so the synthetic tables still join. A
linked key is the join surface. It is aliased consistently rather than
permuted (permuting a key would break the join regardless of masking).

```{r set}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- mask_set(set_dir, mode = "collaborate", seed = 1, quiet = TRUE)
ag <- synthetic(ms)$agronomy
qa <- synthetic(ms)$quality
setequal(unique(ag$gen), unique(qa$gen))   # same genotype codes in both
```

The recipe bundle is private exactly as a single recipe is. `write_set()`
never writes it.

## Geographic coordinates

Latitude and longitude are treated as sensitive: `propose_roles()` flags a
column whose name looks like a coordinate (`gps`, `lat`/`latitude`,
`lon`/`longitude`) as `pii_suspected` and proposes `drop`. Dropping is the
safest choice when the synthetic does not need locations.

**A coordinate is masked unless you say otherwise.** Dropping is one way to
say what happens to it; there are three, and `mask()` refuses to guess. If a
column whose name says coordinate would be written through with
`action = "keep"`, the call stops. Declare the pair to `coords` and it is
coarsened, give the column `drop` or `scramble` and it is masked, or pass
`allow_unmasked_coords = TRUE` having decided the position is not sensitive --
which is recorded on the recipe, so the decision is visible later. A column
that looks like a coordinate only by the shape of its values raises a warning
instead of stopping, because that detection is a guess and a false positive
should not block a legitimate mask.

When the synthetic does need plausible coordinates -- to exercise a spatial
pipeline, say -- a plain scramble is the wrong tool: the copula re-simulates
each axis and smears a latitude/longitude pair into a continuous cloud that can
land in the sea. masque offers two purpose-built alternatives.

`jitter_coordinates()` coarsens coordinates in place by a geographic-masking
jitter. The default donut scheme displaces every point by a random distance
drawn uniformly by area from an annulus, in a random direction, and re-draws
until the point falls on land, so a coastal site is never pushed offshore. The
longitude step is corrected by `cos(latitude)` so the ground distance matches
the requested kilometres at any latitude, and the NA pattern and the
latitude/longitude pairing are preserved.

```{r geo-coords, eval = requireNamespace("maps", quietly = TRUE)}
sites <- data.frame(
  site = c("A", "B", "C"),
  lat  = c(-34.9, -35.2, -33.6),
  lon  = c(138.6, 142.0, 148.2)
)
jitter_coordinates(sites, "lat", "lon", min_km = 5, max_km = 20, seed = 1)
```

**A coordinate belongs to a site, not to a row.** One displacement is drawn per
site and broadcast to every row of that site, so a table holding many rows per
physical place -- plots within a trial, samples within a paddock -- comes back
with one masked coordinate per place, exactly as the source table has one true
coordinate per place. By default a site is a set of rows sharing an identical
input coordinate, which needs no user action and cannot alter genuinely
point-level data. Where a site's recorded coordinates differ slightly between
rows, name the site columns and the group is consolidated to its centroid.

```{r geo-by, eval = requireNamespace("maps", quietly = TRUE)}
plots <- data.frame(
  trial = rep(c("T1", "T2"), each = 4),
  lat   = rep(c(-34.9, -35.2), each = 4),
  lon   = rep(c(138.6, 142.0), each = 4)
)
masked <- jitter_coordinates(plots, "lat", "lon", by = "trial", seed = 1)
nrow(unique(masked[c("lat", "lon")]))   # one coordinate per trial, not per plot
```

Grouping is a confidentiality control, not only a fidelity one. Donut
displacement is isotropic, so the mean of many independent draws around one
true site converges on that site. Displacing a 360-row trial row by row leaves
its centroid a median 0.58 km from the truth, against the 5 km floor the donut
was asked for. One draw per site removes that estimator and leaves the full
displacement in place. Choose the unit carefully: it is the finest grouping
denoting one physical place at one time, so for a multi-year trial series it is
location by year rather than location alone, because trials at one named
location in different seasons legitimately sit in different paddocks.

A site that cannot be placed on land within `max_tries` re-draws is set to `NA`
on both axes, with a warning. masque does not emit a true coordinate inside a
table the caller believes is masked.

Declaring the pair to `mask()` applies the same coarsening as part of masking,
so the coordinates never pass through the copula:

```{r geo-mask, eval = FALSE}
m <- mask(df, roles, mode = "collaborate",
          coords = list(list(lat = "GPS_S", lon = "GPS_E", by = "TRIAL")),
          seed = 1)
```

The recipe records the grouping each declared pair was masked under, so a clone
carries the account of its own coordinate structure.

`synthesise_geospatial()` is the alternative when you would rather re-anchor
points around fake centroids you supply for each region.

A coordinate declared to `coords` is retained in the synthetic, so `audit_mask()`
reports it -- retention is worth knowing about. It reports it as *medium*, noting
that the column was coarsened in place, rather than as the high-leakage
passthrough it would be had the real values been kept.

**How far to displace.** The right magnitude is not a constant. It is calibrated
to the density of the entities being protected, so that the masked point is
spatially k-anonymous -- roughly, at least *k* comparable entities lie closer to
the masked point than the true one (Hampton et al., 2010). Individual-level
urban health data is typically masked with a standard deviation of about one
kilometre, because cities are dense. Fields and farms are far sparser, so
comparable protection needs a much larger displacement; a donut of roughly 5 to
20 km (the default) moves a point across several properties while keeping it in
the same agroclimatic region (Zandbergen, 2014). For a formal guarantee,
calibrate the radii to the local field density rather than relying on the
default.

## Operational guidance

Default to collaborate mode when in doubt. Local mode is for owner-only
development. Treat the recipe as you treat the original data -- same
security class, same access controls. Re-run `audit_mask()` before any
sharing even though collaborate mode runs it for you. Never override a
`pii_suspected` flag without deciding deliberately. Remember that
date/time columns and a preserved design footprint both carry real
operational signal. Alias or roll them up when that signal is sensitive.
Small designs leak -- a categorical cell count of one is high leakage, so
aggregate or drop before masking.

For what the recipe stores and how the round-trip works, see *Recipe
anatomy and the round-trip*.

## References

Hampton, K. H., Fitch, M. K., Allshouse, W. B., Doherty, I. A., Gesink, D. C.,
Leone, P. A., Serre, M. L., & Miller, W. C. (2010). Mapping health data:
improved privacy protection with donut method geomasking. *American Journal of
Epidemiology, 172*(9), 1062--1069. <https://doi.org/10.1093/aje/kwq248>

Zandbergen, P. A. (2014). Ensuring confidentiality of geocoded health data:
assessing geographic masking strategies for individual-level data. *Advances in
Medicine, 2014*, 567049. <https://doi.org/10.1155/2014/567049>
