---
title: "Confidentiality and the threat model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidentiality and the threat model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
library(masque)
set.seed(1)
```

## What masque protects, and what it does not

`masque` is **not** a privacy-preserving or differential-privacy tool.
It is a structurally faithful development surrogate. Read this vignette
before sharing any `masque` output beyond your own machine.

The recipe returned by `mask()` is at least as sensitive as the original
data. The whole design assumes that only the *synthetic* crosses the
trust boundary and the recipe stays with the custodian.

| Actor holds | Wants to learn | What `masque` protects |
|---|---|---|
| Synthetic only | Original raw values | Aliased treatment / categorical vocabularies; jittered numerics; dropped ids and free text; optionally, the column names |
| Synthetic + recipe | Original raw values | Nothing -- the combination is as sensitive as the original |
| Recipe only | Original raw values | Nothing useful -- the recipe is meaningless without the synthetic |
| Synthetic + external side information | Identity of treatments / sites | Only the label vocabulary; a preserved design footprint or a `keep` column is recognisable, and side information wins |

What `masque` does: it preserves enough structure for pipelines to run
unchanged, exposes the privacy-versus-fidelity trade-off through two
explicit modes, records every translation in a private recipe that
round-trips, and audits its own output before you share it.

What `masque` does not do: it gives no differential-privacy guarantee, it
does not make output safe for public release, it does not hide rare
strata, small designs, or operational metadata such as small
site-by-year combinations or contact names, and it does not rewrite
pipeline source code.

## Two axes of control: roles and actions

Since version 0.6.0, every column carries a `role` (what it is) and an
`action` (how deeply it is masked). The mode sets the default action per
role; you override any column you like. The four actions are the privacy
dial:

| Action | Effect | Vocabulary visible? |
|---|---|---|
| `keep` | byte-identical pass-through | yes -- real values published |
| `scramble` | re-simulate numerics; row-permute categoricals / dates | yes -- labels stay, assignment moves |
| `alias` | scramble, then replace labels with opaque codes | no |
| `drop` | column omitted from the synthetic | n/a |

The mode chooses sensible defaults so the common case is safe without
per-column work:

| Knob | `mode = "local"` | `mode = "collaborate"` |
|---|---|---|
| Treatment labels | kept | aliased (`trt_001`) |
| Categorical covariate labels | kept | aliased (`<col>_L001`) |
| Date / time columns | row-permuted; class preserved | row-permuted; class preserved |
| Identifiers (`id`) | kept | dropped |
| Free text (`text`) | kept | dropped |
| Numeric synthesis | empirical-quantile (may emit observed values) | empirical-quantile plus within-resolution jitter; integers stochastically rounded |
| NA mask | preserved cell-by-cell | preserved cell-by-cell |
| `audit_mask()` | on demand | automatic at `mask()` time |
| `print(recipe)` | redacted | redacted; explicit `reveal_maps()` to inspect |

## Depth controls

Two further controls sharpen the depth where the defaults are not enough.

**Hiding column names.** Names themselves can identify a column
(`roseworthy_yield`). `mask(..., alias_names = TRUE)` replaces every
retained column name with an opaque code, which the recipe inverts on the
round-trip. Pass a character vector to hide only some names.

**Hiding design structure.** Design columns are byte-identical by
default, which preserves the experimental layout exactly -- and a
publicly registered trial layout is a fingerprint. To keep the structure
but hide the site or block *labels*, set a design column's action to
`alias`:

```{r design-alias}
df <- data.frame(
  site = factor(rep(c("Roseworthy", "Minnipa", "Turretfield"), each = 6)),
  rep = rep(1:3, 6),
  yield = rnorm(18)
)
roles <- propose_roles(df, detect = FALSE)
roles <- set_role(roles, "site", role = "design", action = "alias")
s <- synthetic(mask(df, roles, mode = "collaborate", seed = 1))
unique(as.character(s$site))   # real site names gone
table(s$site)                  # ... but the three-site structure intact
```

This knowingly breaks byte-identity, so it is never a default -- you ask
for it explicitly.

## The conditional clone: preserving the treatment effect

The default numeric synthesis re-simulates every scrambled numeric column
from a single global Gaussian copula. That preserves each column's
marginal distribution and the global covariance, which is enough to
develop most pipelines -- but it severs the relationship between treatment
and outcome. The synthetic outcomes are drawn from the pooled
distribution and the treatment labels are relabelled independently, so a
causal model fitted on the clone sees no association between an arm and
its response. A pipeline whose whole purpose is to estimate a treatment
effect would silently give the wrong answer on such a clone.

The `conditional = TRUE` argument fixes this. It fits and samples the
copula *within each treatment-by-design stratum* rather than pooling, so a
row's synthetic outcome inherits the location of the treatment that row
carries. The treatment-to-outcome map -- the quantity a causal model
reads -- survives the clone, within sampling tolerance. This is the
data-side analogue of preserving a conditional mean embedding rather than
a pooled marginal.

The contrast is easiest to see on a two-arm trial with a known effect:

```{r conditional-fixture}
set.seed(42)
n <- 600
arm <- factor(rep(c("ctrl", "treat"), each = n / 2))
# A real +5-unit effect of the treated arm on yield.
yield <- 10 + 5 * (arm == "treat") + rnorm(n, sd = 2)
trial <- data.frame(genotype = arm, yield = yield)

roles <- propose_roles(trial, detect = FALSE)
roles <- set_role(roles, "genotype", role = "treatment", action = "scramble")
roles <- set_role(roles, "yield", role = "outcome")

true_effect <- coef(lm(yield ~ genotype, trial))[["genotypetreat"]]
true_effect
```

Cloning both ways from the same seed, then re-estimating the effect on
each clone:

```{r conditional-contrast}
marg <- suppressWarnings(
  mask(trial, roles, mode = "local", seed = 1, conditional = FALSE)
)
cond <- suppressWarnings(
  mask(trial, roles, mode = "local", seed = 1, conditional = TRUE)
)

effect_of <- function(m) {
  coef(lm(yield ~ genotype, synthetic(m)))[["genotypetreat"]]
}

data.frame(
  clone = c("marginal (default)", "conditional"),
  estimated_effect = c(effect_of(marg), effect_of(cond)),
  true_effect = true_effect
)
```

The marginal clone collapses the effect toward zero; the conditional
clone recovers it. Both clones still match the pooled marginal of the
outcome:

```{r conditional-marginal-held}
data.frame(
  source = c("original", "marginal clone", "conditional clone"),
  mean_yield = c(
    mean(trial$yield), mean(synthetic(marg)$yield),
    mean(synthetic(cond)$yield)
  ),
  sd_yield = c(
    sd(trial$yield), sd(synthetic(marg)$yield),
    sd(synthetic(cond)$yield)
  )
)
```

The conditioning columns -- the treatment plus any retained design
columns -- are recorded on the recipe, so the choice is auditable:

```{r conditional-recipe}
recipe(cond)@conditional
recipe(cond)@conditioning_cols
```

Conditional cloning composes with both modes and with `mask_set()` (each
table is stratified by its own treatment and design columns). It needs
enough rows per stratum to fit a stratum-local copula; cells smaller than
a handful of rows are pooled into a graceful global fallback rather than
failing, and with no treatment or design column to condition on the path
degrades cleanly to the global copula with a note. Reach for it whenever
the development pipeline estimates an effect, not just a distribution.

## The leakage audit

`audit_mask()` inspects the synthetic against the original and grades the
leakage of each column. In collaborate mode it runs automatically and
warns at `mask()` time. Here is a fixture built to trip it: a
PII-suspected column the user retains against the default, and a
categorical covariate with a frequency-one level.

```{r leaky-fixture}
df <- data.frame(
  rep = rep(1:3, each = 20),
  variety = factor(rep(paste0("V", 1:6), 10)),
  contact_email = factor(rep(c("a@x", "b@y"), 30)),
  rare_treatment = factor(c(
    "only_one",
    sample(c("alpha", "beta", "gamma"), 59, replace = TRUE)
  )),
  yield = rnorm(60, 5, 1),
  stringsAsFactors = FALSE
)

roles <- propose_roles(df, mode = "collaborate")
roles[, c("col", "role", "action", "pii_suspected")]
```

`contact_email` was auto-flagged `pii_suspected` and set to `drop`. We
override it -- pretending the custodian insists on keeping it -- and
make the rare column a covariate, then mask:

```{r leaky-mask}
roles <- set_role(roles, "yield", role = "outcome")
roles <- set_role(roles, "contact_email", role = "covariate", action = "keep")
roles <- set_role(roles, "rare_treatment", role = "covariate")
m <- mask(df, roles, mode = "collaborate", seed = 1)
audit_mask(m)
```

`contact_email` (real values kept across the trust boundary) and
`rare_treatment` (a frequency-one level) are flagged. `masque` responds
on two channels. At construction time, `mask()` raises a classed warning
(`masque_high_leakage`) and records the findings in `recipe@warnings`
and on the object's audit -- the guided `masque()` flow never silences
it. At write time, the package-managed writers (`masque()`'s `out` and
`write_set()`) refuse to write while a HIGH finding stands: nothing is
written, and the flagged columns are listed with the remedy (re-role,
alias, or drop, then mask again). A custodian who has reviewed the
findings can pass `allow_high = TRUE` to write anyway; the override is
raised as a `masque_high_override` warning and recorded in the recipe's
warnings, so the exception stays auditable.

Beyond that gate the release decision stays with the custodian --
whether a synthetic table is appropriate for a given collaborator,
environment, or jurisdiction depends on context the package cannot see.
`masque` informs that decision; it does not make it.

## Multi-table sets

When several tables share a key, `mask_set()` aliases that key
*identically across all of them* so the synthetic tables still join. A
linked key is the join surface; it is aliased consistently rather than
permuted (permuting a key would break the join regardless of masking).

```{r set}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- mask_set(set_dir, mode = "collaborate", seed = 1, quiet = TRUE)
ag <- synthetic(ms)$agronomy
qa <- synthetic(ms)$quality
setequal(unique(ag$gen), unique(qa$gen))   # same genotype codes in both
```

The recipe bundle is private exactly as a single recipe is; `write_set()`
never writes it.

## Operational guidance

Default to collaborate mode when in doubt; local mode is for owner-only
development. Treat the recipe as you treat the original data -- same
security class, same access controls. Re-run `audit_mask()` before any
sharing even though collaborate mode runs it for you. Never override a
`pii_suspected` flag without deciding deliberately. Remember that
date/time columns and a preserved design footprint both carry real
operational signal; alias or roll them up when that signal is sensitive.
Small designs leak -- a categorical cell count of one is high leakage, so
aggregate or drop before masking.

For what the recipe stores and how the round-trip works, see *Recipe
anatomy and the round-trip*.
