---
title: "What a conditional clone keeps of the design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What a conditional clone keeps of the design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~25 s under rmarkdown::render(); macOS arm64, R 4.6.1, one
     core. Eight designs are masked twenty times each under two ladders; the
     multi-environment trial (1,188 rows) accounts for most of it. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
options(width = 100)
set.seed(1)
has_agridat <- requireNamespace("agridat", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r library}
library(masque)
```

## Why

A biometrician is asked to develop a yield model on a synthetic clone of a
multi-environment variety trial, and asks a fair question before starting:
*which parts of the experiment will still be in the clone?* The allocation,
every design column and the treatment labels, is returned as it was. The
outcomes are re-simulated. The question is what a model fitted on those
re-simulated outcomes will find: the treatment effect, the block effect,
the environment effect, or only their marginal histograms.

This vignette answers it for eight design classes with a public exemplar of
each, all from the `agridat` package. Each is masked as a conditional clone
twenty times, the model that the design calls for is fitted on the original
and on every clone, and one pass rule is applied to all eight. The rule is
stated before the table.

## What

`mask(conditional = TRUE)` re-simulates the numeric block within each
treatment-by-design stratum, and coarsens that stratum by a ladder until
every cell holds at least five rows. Two ladders are available.

The default, `ladder = "hierarchy"`, drops plot coordinates first, then
blocking columns, then environment columns, so a county is never dropped
before the replicates inside it, and it carries the main effect of every
dropped column, and of each pair of dropped blocking or environment
columns, into the clone as an additive shift. Rows that even the
treatment-only rung pools into the fallback keep the treatment means the
same way. A level with a single row is pooled with the other singletons
before any effect is estimated, so no row's own outcome is ever carried. A
term that is noise is shrunk to nothing; a strong one is carried in full.
The recipe names what was kept as the stratum (`conditioning_used`), what
was dropped (`conditioning_dropped`) and what was carried as a shift
(`conditioning_shifted`).

`ladder = "levels"` is the ladder of masque 0.11.1 to 0.12.0: it drops the
column with the most distinct values first and keeps nothing of a dropped
column. It is kept so that a clone made under it can be reproduced.

### The pass rule

For each design, over twenty seeds:

* **Allocation.** Every treatment and design column is byte-identical to
  the original in every clone.
* **NA mask.** The outcome's missing cells are the same cells in every
  clone.
* **Blocking and environment terms.** For each such term the original can
  detect (F at or above 2), the median F over the clones is at least half
  the original's.
* **Treatment.** Over the treatment levels with two or more plots, the
  median correlation between the clone's level means and the original's is
  at least 0.7 when the original detects the treatment (F at or above 2); a
  two-level treatment keeps the same top level in most seeds; and the
  spread of the level means stays within a factor of two of the original
  in every case.

The same rule runs in the package's tests on three seeds, and on twenty
when the tests are run off CRAN.

## Do

### The eight designs

```{r specs, eval = has_agridat}
design_specs <- function() {
  augmented <- agridat::kling.augmented[, c("block", "gen", "tsw", "row", "col")]
  n_gen <- table(augmented$gen)
  augmented$check <- factor(ifelse(
    n_gen[as.character(augmented$gen)] > 1L, as.character(augmented$gen), "new"
  ))
  rcbd <- agridat::besag.elbatan
  rcbd$block <- factor(rcbd$col)
  list(
    CRD = list(
      data = agridat::cochran.crd, outcome = "inf", treatment = "trt",
      design = c("row", "col"),
      formula = inf ~ trt, terms = c(trt = "treatment")
    ),
    RCBD = list(
      data = rcbd, outcome = "yield", treatment = "gen",
      design = c("block", "row", "col"),
      formula = yield ~ block + gen,
      terms = c(block = "block", gen = "treatment")
    ),
    `Latin square` = list(
      data = agridat::fisher.latin, outcome = "yield", treatment = "trt",
      design = c("row", "col"),
      formula = yield ~ factor(row) + factor(col) + trt,
      terms = c(`factor(row)` = "block", `factor(col)` = "block",
        trt = "treatment")
    ),
    `split-plot` = list(
      data = agridat::yates.oats[, c("block", "row", "col", "gen", "nitro", "yield")],
      outcome = "yield", treatment = c("gen", "nitro"),
      design = c("block", "row", "col"),
      formula = yield ~ block + gen * nitro,
      terms = c(block = "block", gen = "treatment", nitro = "treatment")
    ),
    `alpha lattice` = list(
      data = agridat::john.alpha[, c("rep", "block", "gen", "yield", "row", "col")],
      outcome = "yield", treatment = "gen",
      design = c("rep", "block", "row", "col"),
      formula = yield ~ rep + rep:block + gen,
      terms = c(rep = "block", `rep:block` = "block", gen = "treatment")
    ),
    augmented = list(
      data = augmented, outcome = "tsw", treatment = c("gen", "check"),
      design = c("block", "row", "col"),
      formula = tsw ~ block + check,
      terms = c(block = "block", check = "treatment")
    ),
    MET = list(
      data = agridat::besag.met, outcome = "yield", treatment = "gen",
      design = c("county", "rep", "block", "row", "col"),
      formula = yield ~ county + county:rep + gen,
      terms = c(county = "environment", `county:rep` = "block",
        gen = "treatment")
    ),
    `repeated measures` = list(
      data = agridat::kenward.cattle, outcome = "weight", treatment = "trt",
      design = c("animal", "day"),
      formula = weight ~ trt + animal + factor(day),
      terms = c(trt = "treatment", animal = "block", `factor(day)` = "block")
    )
  )
}
```

The augmented design is analysed the classical way: the replicated checks
are fixed levels and every unreplicated entry is one class, `new`. A clone
cannot carry an unreplicated entry's own value, by construction, so that is
the comparison the design allows.

The RCBD exemplar records its three complete blocks as the column of the
field; the block factor is made from it and the column kept as a
coordinate.

### Masking and measuring

The roles are the ones a custodian would set: the outcome, the treatment
columns kept as they are, and every design column kept as it is. Local mode
keeps the treatment labels, so the allocation check is a byte comparison.

```{r runner, eval = has_agridat}
design_roles <- function(spec) {
  r <- propose_roles(spec$data, mode = "local", detect = FALSE)
  r <- set_role(r, spec$outcome, role = "outcome")
  r <- set_role(r, spec$treatment, role = "treatment", action = "keep")
  set_role(r, spec$design, role = "design", action = "keep")
}

level_means <- function(d, col, outcome) {
  tab <- table(d[[col]])
  m <- tapply(d[[outcome]], d[[col]], mean, na.rm = TRUE)
  m[names(tab)[tab >= 2L]]
}

design_run <- function(spec, seeds, ladder = "hierarchy") {
  d <- spec$data
  r <- design_roles(spec)
  f_orig <- anova(lm(spec$formula, data = d))
  terms <- names(spec$terms)
  alloc <- c(spec$treatment, spec$design)
  per_seed <- lapply(seeds, function(s) {
    m <- suppressWarnings(
      mask(d, r, mode = "local", seed = s, conditional = TRUE, ladder = ladder)
    )
    sy <- as.data.frame(synthetic(m))
    f_clone <- anova(lm(spec$formula, data = sy))
    trt <- t(vapply(spec$treatment, function(tc) {
      mo <- level_means(d, tc, spec$outcome)
      ms <- level_means(sy, tc, spec$outcome)[names(mo)]
      c(
        corr = if (length(mo) > 2L) cor(mo, ms) else NA_real_,
        ratio = sd(ms) / sd(mo),
        order = as.numeric(identical(names(which.max(mo)), names(which.max(ms))))
      )
    }, numeric(3)))
    list(
      alloc = all(vapply(alloc, function(cn) identical(sy[[cn]], d[[cn]]), logical(1))),
      na = identical(is.na(sy[[spec$outcome]]), is.na(d[[spec$outcome]])),
      F = setNames(f_clone[terms, "F value"], terms),
      trt = trt,
      used = paste(recipe(m)@conditioning_used, collapse = " x "),
      shifted = paste(recipe(m)@conditioning_shifted, collapse = ", ")
    )
  })
  F_clone <- do.call(rbind, lapply(per_seed, `[[`, "F"))
  trt <- simplify2array(lapply(per_seed, `[[`, "trt"))
  list(
    n = nrow(d),
    F_orig = setNames(f_orig[terms, "F value"], terms),
    F_clone = apply(F_clone, 2, median),
    alloc = all(vapply(per_seed, `[[`, logical(1), "alloc")),
    na = all(vapply(per_seed, `[[`, logical(1), "na")),
    trt_corr = apply(trt[, "corr", , drop = FALSE], 1, median),
    trt_ratio = apply(trt[, "ratio", , drop = FALSE], 1, median),
    trt_order = apply(trt[, "order", , drop = FALSE], 1, mean),
    used = per_seed[[1]]$used,
    shifted = per_seed[[1]]$shifted
  )
}

design_verdict <- function(spec, res, f_floor = 0.5, corr_floor = 0.7) {
  kind <- spec$terms
  block_terms <- names(kind)[kind != "treatment"]
  block_ok <- vapply(block_terms, function(tm) {
    res$F_orig[[tm]] < 2 || res$F_clone[[tm]] >= f_floor * res$F_orig[[tm]]
  }, logical(1))
  trt_detected <- any(res$F_orig[names(kind)[kind == "treatment"]] >= 2)
  trt_ok <- vapply(seq_along(spec$treatment), function(i) {
    corr <- res$trt_corr[[i]]
    signal <- if (is.na(corr)) res$trt_order[[i]] > 0.5 else corr >= corr_floor
    (signal || !trt_detected) && res$trt_ratio[[i]] >= 0.5 && res$trt_ratio[[i]] <= 2
  }, logical(1))
  c(allocation = res$alloc, na_mask = res$na,
    blocking = all(block_ok), treatment = all(trt_ok))
}
```

```{r run, eval = has_agridat}
specs <- design_specs()
seeds <- 1:20
runs <- lapply(specs, design_run, seeds = seeds)
runs_levels <- lapply(specs, design_run, seeds = seeds, ladder = "levels")
```

### The table

```{r table, eval = has_agridat}
verdicts <- t(mapply(design_verdict, specs, runs))
summary_tbl <- data.frame(
  design = names(specs),
  rows = vapply(runs, `[[`, numeric(1), "n"),
  stratum = vapply(runs, `[[`, character(1), "used"),
  carried = vapply(runs, `[[`, character(1), "shifted"),
  allocation = verdicts[, "allocation"],
  na_mask = verdicts[, "na_mask"],
  blocking = verdicts[, "blocking"],
  treatment = verdicts[, "treatment"],
  verdict = ifelse(rowSums(!verdicts) == 0, "pass", "fail"),
  row.names = NULL
)
knitr::kable(summary_tbl)
```

```{r terms, eval = has_agridat}
term_tbl <- do.call(rbind, lapply(names(specs), function(nm) {
  data.frame(
    design = nm,
    term = names(specs[[nm]]$terms),
    kind = unname(specs[[nm]]$terms),
    F_original = round(runs[[nm]]$F_orig, 2),
    F_levels = round(runs_levels[[nm]]$F_clone, 2),
    F_hierarchy = round(runs[[nm]]$F_clone, 2),
    row.names = NULL
  )
}))
knitr::kable(term_tbl)
```

```{r treatment, eval = has_agridat}
trt_tbl <- do.call(rbind, lapply(names(specs), function(nm) {
  data.frame(
    design = nm,
    treatment = specs[[nm]]$treatment,
    corr_levels = round(runs_levels[[nm]]$trt_corr, 2),
    corr_hierarchy = round(runs[[nm]]$trt_corr, 2),
    spread_ratio = round(runs[[nm]]$trt_ratio, 2),
    row.names = NULL
  )
}))
knitr::kable(trt_tbl)
```

### Figure: the F statistic per term, clone against original

```{r figure, eval = has_agridat && has_ggplot2, fig.height = 5.5, fig.cap = "F statistic of each model term on the clone as a share of the original, median over twenty seeds, by ladder. Points on the dashed line are terms the clone reproduces exactly; the levels ladder loses every blocking and environment term, the hierarchy ladder keeps them.", fig.alt = "Dot plot of clone-to-original F ratios for every design term, two ladders side by side."}
plot_df <- rbind(
  data.frame(term_tbl[, c("design", "term", "kind")], ladder = "levels",
    ratio = term_tbl$F_levels / term_tbl$F_original),
  data.frame(term_tbl[, c("design", "term", "kind")], ladder = "hierarchy",
    ratio = term_tbl$F_hierarchy / term_tbl$F_original)
)
plot_df <- plot_df[term_tbl$F_original >= 2, ]
plot_df$label <- paste(plot_df$design, plot_df$term, sep = ": ")
ggplot2::ggplot(plot_df, ggplot2::aes(x = ratio, y = label, colour = ladder)) +
  ggplot2::geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_vline(xintercept = 0.5, linetype = "dotted", colour = "grey50") +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_colour_manual(values = c(levels = "#E69F00", hierarchy = "#0072B2")) +
  ggplot2::facet_grid(kind ~ ., scales = "free_y", space = "free_y") +
  ggplot2::labs(x = "F on the clone / F on the original (log scale)", y = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    strip.text.y = ggplot2::element_text(angle = 0)
  )
```

```{r no-agridat, eval = !has_agridat, echo = FALSE, results = "asis"}
cat("The agridat package is not installed on this build, so the designs",
    "are not masked here; the tables and the figure are omitted.\n")
```

## Read

```{r read-values, eval = has_agridat, include = FALSE}
n_pass <- sum(summary_tbl$verdict == "pass")
met <- runs$MET
met_levels <- runs_levels$MET
oats <- runs$`split-plot`
oats_levels <- runs_levels$`split-plot`
```

`r if (has_agridat) sprintf("All %d of the %d designs pass the rule under the hierarchy ladder.", n_pass, nrow(summary_tbl))`
The allocation and the NA mask come back identical in every clone of every
design, under both ladders; that part of the claim does not depend on the
ladder at all. What the ladder decides is what a model finds.

The multi-environment trial is the case the hierarchy ladder was built
for. Its county effect has an F of
`r if (has_agridat) round(met$F_orig[["county"]])` on the original; the
levels ladder drops `county` before `rep`, and the clone's county F is
`r if (has_agridat) round(met_levels$F_clone[["county"]], 1)`. The
hierarchy ladder conditions on genotype alone and carries `county`,
`rep`, `block` and their pairs as shifts; the clone's county F is
`r if (has_agridat) round(met$F_clone[["county"]])`, the nested
`county:rep` term is at
`r if (has_agridat) round(met$F_clone[["county:rep"]], 1)` against
`r if (has_agridat) round(met$F_orig[["county:rep"]], 1)`, and the
genotype means correlate at
`r if (has_agridat) round(met$trt_corr[["gen"]], 2)` with the original
(`r if (has_agridat) round(met_levels$trt_corr[["gen"]], 2)` under the
levels ladder).

Yates' oats shows the same thing on a single trial: the block F is
`r if (has_agridat) round(oats_levels$F_clone[["block"]], 1)` under the
levels ladder and
`r if (has_agridat) round(oats$F_clone[["block"]], 1)` under the hierarchy
ladder, against `r if (has_agridat) round(oats$F_orig[["block"]], 1)` on
the original, while the nitrogen response is kept by both.

The alpha lattice and the RCBD are three-replicate variety trials, and
there the treatment-only rung is itself below the floor of five rows: every
row is pooled. Under the levels ladder that clone is the marginal clone,
with a genotype correlation near zero; under the hierarchy ladder the
genotype means are carried as a shift and the correlation is
`r if (has_agridat) round(runs[["alpha lattice"]]$trt_corr[["gen"]], 2)` on
the lattice and `r if (has_agridat) round(runs$RCBD$trt_corr[["gen"]], 2)`
on the RCBD.

The augmented design is the one where the rule has the least to say: its
block and check terms are not detectable on the original (F below 2), so
only the allocation, the NA mask and the spread of the check means are
tested. The clone does not, and cannot, carry the unreplicated entries'
values.

## Limits

The rule tests fixed-effects F statistics from `lm()`. A split-plot or a
repeated-measures analysis would normally use a mixed model with the right
error strata; the F ratios here compare the same fixed-effects model on
both sides and say nothing about variance components. The treatment check
compares level means over replicated levels, so it does not see a
treatment-by-environment interaction, which the clone does not preserve:
the copula is fitted within the stratum the ladder reached, and once
`county` has been dropped the genotype-by-county pattern is gone. What the
clone carries of a dropped column is an additive main effect, shrunk toward
zero when the column explains little; a spatial trend recorded only as row
and column numbers is carried as row and column means, not as a surface.
The floors in the rule (half the original F, a correlation of 0.7, a
spread within a factor of two) are the tolerances this vignette states,
not properties of the method; a stricter reader should tighten them and
re-run the chunk.

## What to read next

*Confidentiality and the threat model* explains what carrying a level mean
into the clone means for disclosure, and why `conditional = TRUE` is a
statement about what the clone reproduces. *Getting started with masque*
is the custodian's path from a table to a synthetic and a recipe.

## Reproduce

`set.seed(1)` is set once for the document; every `mask()` call passes its
own `seed`, one to twenty, so each clone is reproducible on its own.
Package versions follow.

```{r session-info}
sessionInfo()
```
