---
title: "Automatic design detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automatic design detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 96
)
set.seed(1)
library(masque)
has_agridat <- requireNamespace("agridat", quietly = TRUE)
```

`masque` ships an experimental-design detector. Given a tabular dataset,
`detect_design()` returns the most likely design class — `CRD`, `RCBD`,
`IBD/alpha-lattice`, `row-column`, `split-plot`, `factorial`, or `none`
— with the per-rule evidence behind that choice.

The detector is read-only. It does not mutate `df`, does not change
`mask()` behaviour, and never auto-corrects anything. It recommends a
role assignment that the user can accept, edit, or ignore.

Why bother? Two reasons. First, **structural awareness**: roles inferred
purely from column names miss treatments that lack a `^trt`-style name
(for example the `gen` column in an alpha-lattice). Second, **sanity
validation**: a one-line `plot()` shows whether what you think is the
design is what's actually in the data — a class of bug field
statisticians have lived with for decades.

## The verb in one line

```{r}
ds <- detect_design(iris)
ds
```

The print method shows the picked class (`CRD`), the working treatment
(`Species`), and the top three alternates so you can see how confident
the call was. The full per-rule score is in `ds@scores`:

```{r}
sort(ds@scores, decreasing = TRUE)
```

`ds@recommended_roles` is what gets overlaid on `propose_roles()` when
detection is on (the default since 0.3.0):

```{r}
ds@recommended_roles
```

## Four worked cases

### CRD — Fisher's `iris`

50 observations per species, no block, no spatial. Detection picks
`Species` as the working treatment.

```{r}
plot(detect_design(iris), df = iris)
```

The left panel is replication-per-treatment; the right is per-column
missingness (0 % in `iris`).

### RCBD — synthetic randomised blocks

Five replicates of a 4-treatment design, every treatment in every block:

```{r}
rcbd <- expand.grid(rep = 1:5, trt = factor(letters[1:4]))
rcbd$yield <- rnorm(nrow(rcbd))
ds_rcbd <- detect_design(rcbd)
ds_rcbd
plot(ds_rcbd, df = rcbd)
```

The replication tile shows `trt x rep`; every cell has count 1 — the
defining RCBD signature.

### Split-plot — `agridat::yates.oats`

The classic 1935 split-plot oat trial: six blocks, three nitrogen levels
(whole-plot), three genotypes (sub-plot).

```{r}
if (has_agridat) {
  oats <- agridat::yates.oats
  ds_oats <- detect_design(oats)
  ds_oats
  plot(ds_oats, df = oats)
}
```

Caveat: the detector cannot fully distinguish a true split-plot from a
factorial-in-blocks, because both have the same data layout. The
whole-plot / sub-plot assignment uses cardinality (fewer levels =
whole-plot), which is a heuristic, not a guarantee. If you know which
factor was actually randomised at the whole-plot level, override
`ds@whole_plot_col` / `ds@sub_plot_col` manually.

### Observational — `mtcars`

`mtcars` has no design at all. Every rule scores below the 0.5
threshold; the verdict is `"none"`.

```{r}
ds_mt <- detect_design(mtcars)
ds_mt
plot(ds_mt, df = mtcars)
```

The plot now degrades gracefully: a fingerprint of treatment-frequency
(empty here — there is no treatment factor) and the per-column NA
pattern.

## Integration with `propose_roles()`

`propose_roles(df)` runs detection by default and folds the recommended
roles into the returned tibble. The detected `design_summary` is
stashed as `attr(roles, "design")` so you can `plot()` it without
re-running the detector:

```{r}
if (has_agridat) {
  roles <- propose_roles(agridat::john.alpha)
  print(attr(roles, "design")@class_label)
  print(roles[
    roles$role %in% c("treatment", "design"),
    c("col", "role", "notes")
  ])
}
```

Without detection (v0.2 behaviour), `gen` would have been left as
`covariate` because its column name does not match any treatment
regex. The structural detector promotes it.

Detection also respects explicit user intent. Columns already roled as
`outcome`, `keep`, or `ignore` are excluded from detector candidates, so
a metadata column marked `keep` will not be promoted into a design hint.

Pass `detect = FALSE` to disable:

```{r}
roles_v2 <- propose_roles(iris, detect = FALSE)
roles_v2$role[roles_v2$col == "Species"]
```

## Caveats and limits

- **Heuristic, not guarantee.** The rule engine matches structural
  signatures; it does not know which factor was randomised at which
  level. Always treat the recommendation as a starting point.
- **Split-plot vs factorial.** Structurally identical when the
  whole-plot and sub-plot factors are both fully crossed within each
  block. The detector picks split-plot only when a design-named block
  factor is present and the smaller-cardinality candidate becomes the
  whole-plot.
- **Small fixtures.** Detection of structure on fewer than 20 rows is
  unreliable. When in doubt, pass `detect = FALSE` and assign roles by
  hand.
- **Optional interactive prompts.** Pass `interactive = TRUE` to be
  asked to disambiguate when the top-two rule scores are within
  `tie_delta`. By default the simpler design wins ties.
- **Detection is read-only.** It never changes `mask()` synthesis
  behaviour. Only `propose_roles()` consumes its output, and only as
  role hints.
