---
title: "Getting started with masque"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with masque}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS arm64, R 4.5.2, one
     core. No sampler runs in this vignette, so the time is dominated by
     knitr overhead and the small copula fits. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
options(width = 100)
set.seed(1)
```

```{r library}
library(masque)
```

## Why

A field agronomist holds a confidential 72-plot trial and wants an outside
statistician to build a spatial yield model against it, but the
statistician must never see the real plot layout or genotype identities.
The agronomist's question is practical, not abstract: *can I hand over
something that behaves like my trial well enough for someone to develop a
working pipeline on, and get that pipeline back running unmodified on my
real data when they are done?* This vignette walks the custodian's side of
that handoff end to end, on the classical John and Williams (1995)
alpha-design trial shipped with the package, and answers the question
directly by the end.

`masque` is not an anonymiser, and the synthetic table it produces is not a
public-release-safe artefact -- it is a development surrogate, meant to
cross a trust boundary of one: the custodian's collaborator. The companion
vignette *Confidentiality and the threat model* sets out exactly what is
and is not protected. Read it before sharing any output.

## What

`masque()` is the front door: point it at a table (or a folder or
workbook of related tables) and it reads the data, proposes a masking
plan, masks, and -- in an interactive session -- pauses to let the
custodian review the plan first.

The output of `masque()` is always a pair: the *synthetic* clone, the only
part that crosses the trust boundary, and a private `masque_recipe` (see
`recipe()`) that never leaves the custodian's machine. The recipe is what
lets a pipeline written against the synthetic re-target to the original
later with no code change; its internals and the round-trip verbs are the
subject of *Recipe anatomy and the round-trip*.

Every column in the masking plan carries two independent decisions, built
by `propose_roles()` and adjusted with `set_role()`: a `role` (what the
column is -- `design`, `treatment`, `outcome`, `covariate`, `date`, `id`,
`text`, or `other`) and an `action` (how deeply it is masked -- `keep`
passes a column through byte-for-byte, `scramble` re-simulates numerics
through a Gaussian copula or row-permutes categoricals and dates, `alias`
scrambles and then replaces the labels with opaque codes, and `drop`
leaves the column out entirely). `role_options()` renders the full grid
the validator accepts, so an incompatible pairing -- a design column
`masque` refuses to scramble, for instance -- can be checked before it is
attempted.

The `mode` argument sets the safe defaults for both axes at once: `local`
is for the custodian's own development, and keeps vocabulary intact;
`collaborate` is for handing the synthetic to someone else, and aliases
treatment and categorical labels, jitters numerics, drops identifiers and
free text, and runs the leakage audit automatically. Per-column `action`
choices always override the mode's default.

## Do

### The one-call path

```{r one-call}
f <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
knitr::kable(
  head(df),
  caption = "The first six plots of the alpha-design trial."
)

m <- masque(df, mode = "collaborate", seed = 1L, ask = FALSE)
```

`ask = FALSE` skips the interactive review, which is what a vignette
needs; in an interactive console, `masque(df)` shows the proposed plan and
a prompt to proceed, edit, or stop. The message above is the summary
`masque()` prints for every guided run: how many columns were masked, in
which mode, and the leakage-audit tally.

The result carries the synthetic data and the private recipe:

```{r one-call-extract}
synth <- synthetic(m)
knitr::kable(
  head(synth),
  caption = paste0(
    "The first six plots of the synthetic clone: `gen` is aliased, ",
    "`yield` is re-simulated."
  )
)
```

### The masking plan: roles and actions

Under the one-call path, `masque()` built the roles table without asking.
Building and reviewing it directly gives full control:

```{r roles}
roles <- propose_roles(df, mode = "collaborate")
knitr::kable(
  roles[, c("col", "role", "action", "kind")],
  caption = paste0(
    "The proposed masking plan for the alpha-design trial ",
    "(collaborate mode)."
  )
)
```

`propose_roles()` fills in a sensible action for each column given the
mode, so the table reviewed here is the plan that will run. `yield` was
proposed as a `covariate`; naming it the trial's `outcome` changes nothing
about how it is masked, but it documents intent and is what a downstream
consumer such as the *conditional* clone (see *Confidentiality and the
threat model*) reads to find the response:

```{r set-role}
roles <- set_role(roles, "yield", role = "outcome")
knitr::kable(
  roles[, c("col", "role", "action")],
  caption = "The plan after naming `yield` as the trial outcome."
)
```

Re-assigning a role re-resolves the default action; passing an explicit
`action` pins the column instead. There is no requirement to name an
outcome: with none marked, every scrambled numeric is re-simulated jointly.

### Editing the plan as code

The printed table -- and the spreadsheet the guided prompt opens when the
custodian chooses `e` -- is an ordinary data frame, so anything the editor
can do, a script can do reproducibly. `set_role()` is vectorised over
columns:

```{r edit-vectorised}
r2 <- set_role(roles, c("row", "col"), action = "drop")
```

Direct edits work too. A direct `role` edit leaves `action` untouched, so
setting the action back to `NA` tells `mask()` to re-resolve the default
for the new role:

```{r edit-direct}
r2$role[r2$col == "rep"] <- "covariate"
r2$action[r2$col == "rep"] <- NA
knitr::kable(
  r2[, c("col", "role", "action")],
  caption = paste0(
    "A hand-edited plan: `row`/`col` dropped, `rep` re-roled to a ",
    "covariate."
  )
)
```

The `kind` column is derived from the column's class, never chosen;
editing it changes nothing, and the column should be converted in the data
and re-proposed if the kind is wrong.

`role_options()` renders the full grid the validator accepts, filtered by
a column's storage kind:

```{r role-options}
knitr::kable(
  role_options(kind = "factor"),
  caption = paste0(
    "Every (role, action) pair the validator accepts for a factor ",
    "column."
  )
)
```

Finally, `pii_suspected`. `propose_roles()` sets it from the column
*name* (`email`, `phone`, `owner`, ...), and the leakage audit treats a
flagged column that survives into the synthetic as a HIGH finding. The
scan reads names, not content, so when a harmlessly named column holds
sensitive values, flagging it by hand is the correct fix:

```{r pii-by-hand, eval = FALSE}
roles$pii_suspected[roles$col == "comments"] <- TRUE
```

Passing the edited table back to `mask()` (or to `masque(df, roles =
roles)`) runs the plan:

```{r mask}
m <- mask(df, roles, mode = "collaborate", seed = 1L)
synth <- synthetic(m)
byte_identical_plot <- identical(synth$plot, df$plot)
gen_levels_changed <- !setequal(levels(synth$gen), levels(df$gen))
byte_identical_plot
gen_levels_changed
head(levels(synth$gen))
```

### A refusal you can rely on

Not every role and action pairing makes sense -- `plot` is a `design`
column, and a design column is structure rather than content, so asking
`mask()` to scramble it is a request the package will not carry out:

```{r fail-closed, error = TRUE}
bad_roles <- set_role(roles, "plot", role = "design", action = "scramble")
mask(df, bad_roles, mode = "collaborate", seed = 1L)
```

The call aborts before any masking happens, naming the exact column and
combination at fault and offering the three actions that do make sense for
a design column (`keep`, `alias`, `drop`). Nothing is silently
reinterpreted -- an invalid plan is a refusal, not a best-effort guess, and
the same validation runs whether the plan came from `propose_roles()`,
hand-editing, or the guided spreadsheet flow.

### More than one table

A confidential dataset often arrives as several related files or a
multi-sheet workbook. `masque()` handles those too -- point it at a
folder, an `.xlsx` file, or a named list of data frames and it masks every
table at once, aliasing any shared key (a site code, a genotype name)
*identically across tables* so a join of the synthetic tables still
resolves.

```{r set}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- masque(set_dir, mode = "collaborate", seed = 1L, ask = FALSE)
ms
```

The genotype column `gen` appears in both tables and is masked to the same
codes in each, so the field and laboratory tables still join. See
*Confidentiality and the threat model* for the set-level controls.

### Multi-environment structure

A multi-environment trial has at least three distinct structural
questions: which rows belong to each environment, whether treatments
connect the environments, and what randomisation structure can be
recovered within each environment. `detect_design()` reports these
separately; connectivity is a comparability diagnostic, not the definition
of a multi-environment trial, and an observed block or field layout is
evidence, not proof, of the original randomisation protocol.

A small synthetic toy with two environments and three genotypes makes the
three questions concrete:

```{r met-scope}
met <- expand.grid(
  env = factor(c("E1", "E2")),
  rep = factor(seq_len(2L)),
  gen = factor(c("G1", "G2", "G3"))
)
met$yield <- seq_len(nrow(met)) + rep(c(0, 2), each = 6L)

ds <- detect_design(met)
ds@scope_label
ds@env_cols
ds@connectivity$status
ds@within_design_label
```

Automatic detection is deliberately conservative: exact `env`/`environment`
names and bounded site-year patterns are selected only when they pass
validity and competition gates, and a site-only column auto-resolves only
when treatments are replicated across sites, so that a nested block is not
promoted into an environment by mistake. Supplying the basis explicitly is
the right call when domain knowledge is stronger than the recorded names:

```{r met-explicit}
ds_explicit <- detect_design(met, env = "env")
```

### Figure: the multi-environment coverage plot

`plot.design_summary()` draws a compact environment overview by default,
in base graphics or, with `engine = "ggplot2"`, as a `ggplot2` object --
this is the figure below. It shows something the two console values above
cannot: whether coverage is *even* across environments, which a bare
connectivity flag does not distinguish from one environment barely
scraping in.

```{r met-plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.cap = "Genotypes observed per environment (E1, E2) for the toy multi-environment trial; the subtitle carries connectivity status, component count, and the recovered within-environment design."}
plot(ds_explicit, df = met, engine = "ggplot2")
```

```{r met-plot-skip, eval = !requireNamespace("ggplot2", quietly = TRUE), echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the coverage figure is",
    "skipped; `plot(ds_explicit, df = met)` still draws the base-graphics",
    "equivalent.\n")
```

Both bars sit at three treatments observed, because every genotype in this
toy trial appears in both environments -- that even coverage, together
with the subtitle's connected status and single component, is what makes
the two environments comparable at all. A design where a genotype fails to
show up in one bar, or where the subtitle reports more than one component,
marks environments that cannot be pooled into a single connected analysis
without first checking whether the missing coverage is real or an artefact
of how the table was assembled.

High-confidence environment recommendations feed the masking plan directly.
Local mode keeps environment values byte-identical; collaborate mode
aliases categorical environment labels in place, preserving row
assignment, factor codes, the NA mask, and recipe inversion. A numeric
environment such as year remains `keep` and raises a disclosure warning
because its values are still visible:

```{r met-roles}
met_roles <- propose_roles(met, mode = "collaborate")
knitr::kable(
  met_roles[met_roles$col == "env", c("col", "role", "action")],
  caption = "How the masking plan treats the `env` column."
)
```

This safeguard protects the allocation a pipeline reads. It does not imply
that the synthesised outcomes preserve genotype-by-environment effects:
sparse treatment-by-environment cells may fall back to pooled synthesis,
and the clone must not be used as a substitute for the original trial in
scientific inference.

## Read

The one-call run masked all seven columns of the alpha-design trial in
collaborate mode with an audit tally of zero HIGH and zero medium
findings, so nothing about this particular trial needed the custodian's
attention before the synthetic could be handed over. On the hand-edited
plan, `plot` -- a `design` column left at `keep` -- came back
`r isTRUE(byte_identical_plot)` for byte-identity against the original,
while `gen` -- a `treatment` column aliased -- came back with its levels
changed (`r isTRUE(gen_levels_changed)`): the synthetic vocabulary is
`trt_001` through `trt_024`, not the original genotype codes, exactly the
two-tier behaviour the roles table promised. The refusal above shows the
same plan being checked before it runs: asking to scramble a design column
is rejected outright, not silently downgraded to something that would run.

The multi-environment toy resolves to `r ds@connectivity$status`
connectivity with `r ds@connectivity$components` component and a within-
environment design of `r ds@within_design_label`, which the figure repeats
visually as two equal-height bars -- the numbers and the picture agree
because every genotype in this toy happens to appear in both environments.
The `met_set` fixture masks two related tables at once and keeps `env` and
`gen` aliased identically across both, which is what lets the field and
laboratory tables still join after masking.

Answering the opening question directly: yes, a custodian can hand over a
synthetic clone that a collaborator can develop a full pipeline against,
while the plot layout, genotype identities, and any linked tables stay
tied together exactly as they were, aliased rather than exposed. The other
half of that handoff -- how the collaborator's pipeline gets back onto the
real data -- is *Recipe anatomy and the round-trip*.

## Limits

This vignette shows the custodian's side of the handoff, on one small
public trial with a clean, complete design; it does not exercise the
leakage audit's failure modes, the geographic-coordinate controls, or the
conditional clone that preserves a treatment effect rather than only a
marginal distribution, all of which live in *Confidentiality and the
threat model*. `pii_suspected` detection reads column names, not content,
so a harmlessly named column holding sensitive values is caught only if
someone flags it by hand, as shown above -- the package cannot infer
sensitivity it has no textual evidence for. Multi-environment detection is
deliberately conservative and can under-detect a genuine environment
structure when names and replication patterns are ambiguous; a domain
expert's explicit `env` argument should be preferred over the automatic
guess whenever the two disagree. Finally, this vignette never scrambles or
aliases a numeric column jointly with others, so it does not show what the
Gaussian copula does or does not preserve about the relationship between
columns -- that is also *Confidentiality and the threat model*'s subject.

## What to read next

*Confidentiality and the threat model* sets out exactly what a synthetic
clone protects and what it does not, the depth controls beyond role and
action, the leakage audit, and the conditional clone that preserves a
treatment-to-outcome relationship. *Recipe anatomy and the round-trip*
is the analyst's side of this handoff: what the recipe carries, and how a
pipeline built on the synthetic re-targets to the original.

## Reproduce

`set.seed(1)` is set once for the whole document; every `masque()`/`mask()`
call also passes `seed = 1L` explicitly, so each is independently
reproducible regardless of what ran before it in this vignette. Package
versions follow.

```{r session-info}
sessionInfo()
```
