---
title: "Getting started with masque"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with masque}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
set.seed(1)
```

## What masque is for

`masque` exists to bridge an expertise gap. A data custodian holds a
confidential dataset and the legal responsibility for it, but often only
basic R fluency. An analyst has the modelling expertise but cannot
lawfully see the raw data. `masque` lets the custodian hand the analyst a
structurally faithful *synthetic* clone -- close enough that a pipeline
developed against it runs unchanged on the real data -- while the
information that must stay private never crosses the boundary.

It turns one tabular dataset (or a folder or workbook of related tables)
into a synthetic clone whose experimental design, missing-value pattern,
and global covariance are preserved, alongside a private `recipe` that
round-trips: a pipeline written against the synthetic re-targets to the
original with no code changes.

It is **not** an anonymiser. The synthetic is a development surrogate,
not a public-release-safe artefact. The companion vignette
*Confidentiality and the threat model* sets out exactly what is and is
not protected; read it before sharing any output.

## The one-call path

`masque()` is the front door. Point it at your data and it reads the
table, proposes a masking plan, masks, and -- in an interactive session
-- pauses to let you review the plan first. We use the classical John &
Williams (1995) alpha-design field trial, shipped with the package.

```{r one-call}
library(masque)

f <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
head(df)

m <- masque(df, mode = "collaborate", seed = 1, ask = FALSE)
```

`ask = FALSE` skips the interactive review, which is what we want inside
a vignette. In your own console, call `masque(df)` and you will see the
proposed plan and a prompt to proceed, edit, or stop.

The result carries the synthetic data and the private recipe:

```{r one-call-extract}
synth <- synthetic(m)
head(synth)
```

## The masking plan: roles and actions

Under the one-call path, `masque()` builds the *roles table* for you. You
can build and edit it yourself for full control. Every column gets two
decisions: a `role` (what the column is) and an `action` (what masque
does to it).

```{r roles}
roles <- propose_roles(df, mode = "collaborate")
roles[, c("col", "role", "action", "kind")]
```

The eight roles -- `design`, `treatment`, `outcome`, `covariate`,
`date`, `id`, `text`, `other` -- describe the column. The four actions
set the depth:

- `keep` -- pass the column through byte-for-byte;
- `scramble` -- re-simulate numerics through a Gaussian copula, or
  row-permute categoricals, dates, and text (the vocabulary stays
  visible);
- `alias` -- scramble and replace the labels with opaque codes;
- `drop` -- leave the column out of the synthetic entirely.

`propose_roles()` fills in a sensible action for each column given the
mode, so the table you review is the plan that will run. Edit it with
`set_role()`:

```{r set-role}
roles <- set_role(roles, "yield", role = "outcome")
roles[, c("col", "role", "action")]
```

Re-assigning a role re-resolves the default action; passing an explicit
`action` pins the column. There is no requirement to name an `outcome`:
with none marked, every scrambled numeric is re-simulated jointly.

Pass the edited table back to `mask()` (or to `masque(df, roles =
roles)`):

```{r mask}
m <- mask(df, roles, mode = "collaborate", seed = 1)
synth <- synthetic(m)
identical(synth$plot, df$plot)               # design column: byte-identical
setequal(levels(synth$gen), levels(df$gen))  # treatment: aliased away
head(levels(synth$gen))
```

## Local and collaborate modes

The mode sets the safe defaults.

| Mode | Use case | Defaults |
|---|---|---|
| `local` | Owner develops on a realistic surrogate locally | Vocabulary preserved; numeric values may match observed |
| `collaborate` | Owner shares the synthetic while keeping the recipe private | Treatment and categorical labels aliased; numerics jittered; ids and free text dropped; the leakage audit runs automatically |

Per-column `action` choices override the mode wherever you need them.

## Tidy, dates, and depth

Real custodian tables are rarely clean. `masque()` legalises column
names and trims stray whitespace before masking, reports near-duplicate
labels (likely typos) without merging them, and records every fix so the
round-trip still lines up. Set `clean = "report"` to preview the fixes,
or `clean = "off"` to skip them.

Date and time columns get the first-class `date` role: they are
row-permuted, keep their class, and preserve the NA pattern. When even
the column *names* are sensitive, `alias_names = TRUE` hides them behind
opaque codes that the recipe inverts.

## More than one table

A confidential dataset often arrives as several related files or a
multi-sheet workbook. `masque()` handles those too -- point it at a
folder, an `.xlsx` file, or a named list of data frames and it masks
every table at once, aliasing any shared key (a site code, a genotype
name) *identically across tables* so a join of the synthetic tables still
resolves.

```{r set}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- masque(set_dir, mode = "collaborate", seed = 1, ask = FALSE)
ms
```

The genotype column `gen` appears in both tables and is masked to the
same codes in each, so the field and laboratory tables still join. See
*Confidentiality and the threat model* for the set-level controls.

## Where to go next

- *Confidentiality and the threat model* -- what is and is not protected,
  the two modes, the depth controls, and the leakage audit.
- *Recipe anatomy and the round-trip* -- the analyst's side: how a
  pipeline built on the synthetic re-targets to the original.
