---
title: "Getting started with masque"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with masque}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
set.seed(1)
```

## What masque is for

`masque` exists to bridge an expertise gap. A data custodian holds a
confidential dataset and the legal responsibility for it, but often only
basic R fluency. An analyst has the modelling expertise but cannot
lawfully see the raw data. `masque` lets the custodian hand the analyst a
structurally faithful *synthetic* clone -- close enough that a pipeline
developed against it runs unchanged on the real data -- while the
information that must stay private never crosses the boundary.

It turns one tabular dataset (or a folder or workbook of related tables)
into a synthetic clone whose experimental design, missing-value pattern,
and global covariance are preserved, alongside a private `recipe` that
round-trips: a pipeline written against the synthetic re-targets to the
original with no code changes.

It is **not** an anonymiser. The synthetic is a development surrogate,
not a public-release-safe artefact. The companion vignette
*Confidentiality and the threat model* sets out exactly what is and is
not protected. Read it before sharing any output.

## The one-call path

`masque()` is the front door. Point it at your data and it reads the
table, proposes a masking plan, masks, and -- in an interactive session
-- pauses to let you review the plan first. We use the classical John &
Williams (1995) alpha-design field trial, shipped with the package.

```{r one-call}
library(masque)

f <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
head(df)

m <- masque(df, mode = "collaborate", seed = 1, ask = FALSE)
```

`ask = FALSE` skips the interactive review, which is what we want inside
a vignette. In your own console, call `masque(df)` and you will see the
proposed plan and a prompt to proceed, edit, or stop.

The result carries the synthetic data and the private recipe:

```{r one-call-extract}
synth <- synthetic(m)
head(synth)
```

## The masking plan: roles and actions

Under the one-call path, `masque()` builds the *roles table* for you. You
can build and edit it yourself for full control. Every column gets two
decisions: a `role` (what the column is) and an `action` (what masque
does to it).

```{r roles}
roles <- propose_roles(df, mode = "collaborate")
roles[, c("col", "role", "action", "kind")]
```

The eight roles -- `design`, `treatment`, `outcome`, `covariate`,
`date`, `id`, `text`, `other` -- describe the column. The four actions
set the depth:

- `keep` -- pass the column through byte-for-byte.
- `scramble` -- re-simulate numerics through a Gaussian copula, or
  row-permute categoricals, dates, and text (the vocabulary stays
  visible).
- `alias` -- scramble and replace the labels with opaque codes.
- `drop` -- leave the column out of the synthetic entirely.

`propose_roles()` fills in a sensible action for each column given the
mode, so the table you review is the plan that will run. Edit it with
`set_role()`:

```{r set-role}
roles <- set_role(roles, "yield", role = "outcome")
roles[, c("col", "role", "action")]
```

Re-assigning a role re-resolves the default action. Passing an explicit
`action` pins the column. There is no requirement to name an `outcome`:
with none marked, every scrambled numeric is re-simulated jointly.

### Editing the plan as code

The printed table -- and the spreadsheet the guided prompt opens when
you choose `e` -- is an ordinary data frame, so anything the editor can
do, a script can do reproducibly. `set_role()` is vectorised over
columns:

```{r edit-vectorised}
r2 <- set_role(roles, c("row", "col"), action = "drop")
```

Direct edits work too. A direct `role` edit leaves `action` untouched,
so set the action to `NA` when you want `mask()` to re-resolve the
default for the new role:

```{r edit-direct}
r2$role[r2$col == "rep"] <- "covariate"
r2$action[r2$col == "rep"] <- NA
r2[, c("col", "role", "action")]
```

The `kind` column is derived from the column's class, never chosen, and
editing it changes nothing. Convert the column in the data and
re-propose if the kind is wrong.

Not every role and action pair makes sense -- a design column cannot be
scrambled, an outcome has no labels to alias. `role_options()` renders
the full grid the validator accepts, and its `kind` argument filters it
to what is available for one column's storage kind:

```{r role-options}
role_options(kind = "factor")
```

Finally, `pii_suspected`. `propose_roles()` sets it from the column
*name* (`email`, `phone`, `owner`, ...), and the leakage audit treats a
flagged column that survives into the synthetic as a HIGH finding. The
scan reads names, not content, so when a harmlessly named column holds
sensitive values, flag it yourself and the audit honours the flag:

```{r pii-by-hand, eval = FALSE}
roles$pii_suspected[roles$col == "comments"] <- TRUE
```

Pass the edited table back to `mask()` (or to `masque(df, roles =
roles)`):

```{r mask}
m <- mask(df, roles, mode = "collaborate", seed = 1)
synth <- synthetic(m)
identical(synth$plot, df$plot)               # design column: byte-identical
setequal(levels(synth$gen), levels(df$gen))  # treatment: aliased away
head(levels(synth$gen))
```

## Local and collaborate modes

The mode sets the safe defaults.

| Mode | Use case | Defaults |
|---|---|---|
| `local` | Owner develops on a realistic surrogate locally | Vocabulary preserved, with numeric values that may match the observed values |
| `collaborate` | Owner shares the synthetic while keeping the recipe private | Treatment and categorical labels aliased, numerics jittered, ids and free text dropped, and the leakage audit run automatically |

Per-column `action` choices override the mode wherever you need them.
When `mask()` or `mask_set()` is called without `mode`, it inherits the mode
stored on the reviewed roles plan. Passing `mode = "local"` explicitly with a
collaborate-mode plan raises a warning so a sharing-oriented plan cannot be
silently downgraded.

## Tidy, dates, and depth

Real custodian tables are rarely clean. `masque()` legalises column
names and trims stray whitespace before masking, reports near-duplicate
labels (likely typos) without merging them, and records every fix so the
round-trip still lines up. Set `clean = "report"` to preview the fixes,
or `clean = "off"` to skip them.

Date and time columns get the first-class `date` role: they are
row-permuted, keep their class, and preserve the NA pattern. When even
the column *names* are sensitive, `alias_names = TRUE` hides them behind
opaque codes that the recipe inverts.

## More than one table

A confidential dataset often arrives as several related files or a
multi-sheet workbook. `masque()` handles those too -- point it at a
folder, an `.xlsx` file, or a named list of data frames and it masks
every table at once, aliasing any shared key (a site code, a genotype
name) *identically across tables* so a join of the synthetic tables still
resolves.

```{r set}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- masque(set_dir, mode = "collaborate", seed = 1, ask = FALSE)
ms
```

The genotype column `gen` appears in both tables and is masked to the
same codes in each, so the field and laboratory tables still join. See
*Confidentiality and the threat model* for the set-level controls.

## Multi-environment structure

A multi-environment trial has at least three distinct structural questions:
(1) which rows belong to each environment, (2) whether treatments connect the
environments, and (3) what randomisation structure can be recovered within
each environment. `detect_design()` reports these separately. Connectivity is
a comparability diagnostic, not the definition of a multi-environment trial.
An observed block or field layout is evidence, not proof of the original
randomisation protocol.

This small example has two environments and three genotypes:

```{r met-scope}
met <- expand.grid(
  env = factor(c("E1", "E2")),
  rep = factor(seq_len(2L)),
  gen = factor(c("G1", "G2", "G3"))
)
met$yield <- seq_len(nrow(met)) + rep(c(0, 2), each = 6L)

ds <- detect_design(met)
ds@scope_label
ds@env_cols
ds@connectivity$status
ds@within_design_label
```

Automatic detection is deliberately conservative. Exact `env` or
`environment` names and bounded site-year patterns can be selected when they
pass validity and competition gates. A site-only column auto-resolves only
when treatments are replicated across sites, preventing a nested block from
being promoted as an environment. Weak or competing evidence remains
uncertain. Supply the basis explicitly when domain knowledge is stronger than
the recorded names:

```{r met-explicit}
ds_explicit <- detect_design(met, env = "env")
```

The default plot is a compact environment overview. Select one label to inspect
its recoverable field structure. The original data frame remains an explicit
argument because `design_summary` does not duplicate source data.

```{r met-plot, fig.width = 7, fig.height = 4, fig.cap = "One bar per environment (E1, E2), height the number of genotypes observed there; the subtitle reports connectivity status, the count of connected components, and the recovered within-environment design (here, RCBD).",fig.alt = "Bar chart titled Multi-environment coverage: 2 environments. Two equal-height bars, one for E1 and one for E2, each at three treatments observed. Subtitle reads connected connectivity, 1 component, inner design RCBD."}
plot(ds_explicit, df = met)
```

Both bars sit at three treatments observed, because every genotype in this
toy trial appears in both environments -- that even coverage, together with
the subtitle's "connected connectivity, 1 component", is what makes the two
environments comparable at all. A design where a genotype fails to show up
in one bar, or where the subtitle reports more than one component, marks
environments that cannot be pooled into a single connected analysis without
first checking whether the missing coverage is real or an artefact of how
the table was assembled.

High-confidence environment recommendations feed the masking plan. Local mode
keeps environment values byte-identical. Collaborate mode aliases categorical
environment labels in place, preserving row assignment, factor codes, the NA
mask, and recipe inversion. A numeric environment such as year remains `keep`
and raises a disclosure warning because its values are still visible.

```{r met-roles}
met_roles <- propose_roles(met, mode = "collaborate")
met_roles[met_roles$col == "env", c("col", "role", "action")]
```

This safeguard protects the allocation used by a pipeline. It does not imply
that the synthesised outcomes preserve genotype-by-environment effects. Sparse
treatment-by-environment cells may fall back to pooled synthesis, and the clone
must not be used as a substitute for the original trial in scientific
inference.

## Where to go next

- *Confidentiality and the threat model* -- what is and is not protected,
  the two modes, the depth controls, and the leakage audit.
- *Recipe anatomy and the round-trip* -- the analyst's side: how a
  pipeline built on the synthetic re-targets to the original.
