---
title: "Getting started with masque"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with masque}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(masque)
```

# masque: development surrogates for tabular data

This vignette is a scaffold. The full getting-started walk-through lands in
build-order step 8 (see `plan/masque_v0.2_plan.md` in the development
workspace). For now, it documents the contract that the package promises.

## Threat model (read first)

`masque` is a structurally faithful development surrogate, **not** an
anonymiser. The recipe is at least as sensitive as the original data; never
share the recipe and the synthetic together. See the future
`vignette("confidentiality", package = "masque")` for the full threat model.

## Two modes

| Knob | `mode = "local"` | `mode = "collaborate"` |
|---|---|---|
| Column names | preserved | preserved by default; user may opt-in to aliasing |
| Treatment levels | preserved | opaque aliased (`trt_001`, ...) |
| Categorical covariate levels | preserved | opaque aliased (`<col>_L01`, ...) |
| `ignore` columns | pass-through option | dropped |
| NA mask | preserved cell-by-cell | preserved with explicit warning |
| Numeric synthesis | empirical-quantile | empirical-quantile + within-resolution jitter; integers stochastically rounded |
| `audit_mask()` | on demand | automatic at `mask()` time |

## Workflow

```r
# Once the API lands:
roles  <- propose_roles(my_data)
# (user edits roles)
m      <- mask(my_data, roles, mode = "collaborate", seed = 20260516)

dev    <- synthetic(m)   # share this with the pipeline developer
rec    <- recipe(m)      # private; never shared
```

The collaborator builds their pipeline against `dev`. The owner runs it on
the original through `apply_recipe()` and `unmask()`:

```r
orig_in_synth_space <- apply_recipe(my_data, rec)
preds_synth         <- pipeline(orig_in_synth_space)
preds_orig          <- unmask(preds_synth, rec)
```
