---
title: "Getting started with masque"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with masque}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS arm64, R 4.5.2, one
     core. No sampler runs in this vignette, so the time is dominated by
     knitr overhead and the small copula fits. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
options(width = 100)
set.seed(1)
```

```{r library}
library(masque)
```

## Why

A field agronomist has a confidential 72-plot trial and wants an outside
statistician to build a spatial yield model on it, without the statistician
seeing the real plot layout or genotype identities. This vignette shows the
agronomist's side of that handoff on the John and Williams (1995)
alpha-design trial shipped with the package.

The synthetic table `masque` produces is a development surrogate for one
named collaborator. It is not an anonymised dataset and is not safe for
public release. *Confidentiality and the threat model* sets out what is and
is not protected; read it before sharing any output.

## What

`masque()` is the front door: point it at a table (or a folder or
workbook of related tables) and it reads the data, proposes a masking
plan, masks, and -- in an interactive session -- pauses to let the
custodian review the plan first.

The output of `masque()` is always a pair: the *synthetic* clone, the only
part that crosses the trust boundary, and a private `masque_recipe` (see
`recipe()`) that never leaves the custodian's machine. The recipe is what
lets a pipeline written against the synthetic re-target to the original
later with no code change; its internals and the round-trip verbs are the
subject of *Recipe anatomy and the round-trip*.

Every column in the masking plan has a `role` and an `action`, proposed by
`propose_roles()` and changed with `set_role()`. The role says what the
column is: `design`, `treatment`, `outcome`, `covariate`, `date`, `id`,
`text` or `other`. The action says what happens to it:

* `keep` passes it through unchanged;
* `scramble` re-simulates numerics through a Gaussian copula, or shuffles
  categories and dates between rows;
* `alias` scrambles and then replaces the labels with opaque codes;
* `drop` leaves it out.

`role_options()` lists every role and action pair the package accepts.

The `mode` argument sets the defaults for both axes at once:

* `local` is for the custodian's own development and keeps labels intact.
* `collaborate` is for handing the synthetic to someone else. It aliases
  treatment and categorical labels, jitters numerics, drops identifiers and
  free text, and runs the leakage audit.

A per-column `action` always overrides the mode's default.

## Do

### The one-call path

```{r one-call}
f <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
knitr::kable(
  head(df),
  caption = "The first six plots of the alpha-design trial."
)

m <- masque(df, mode = "collaborate", seed = 1L, ask = FALSE)
```

`ask = FALSE` skips the interactive review, which is what a vignette
needs; in an interactive console, `masque(df)` shows the proposed plan and
a prompt to proceed, edit, or stop. The message above is the summary
`masque()` prints for every guided run: how many columns were masked, in
which mode, and the leakage-audit tally.

The result carries the synthetic data and the private recipe:

```{r one-call-extract}
synth <- synthetic(m)
knitr::kable(
  head(synth),
  caption = paste0(
    "The first six plots of the synthetic clone: `gen` is aliased, ",
    "`yield` is re-simulated."
  )
)
```

### The masking plan: roles and actions

Under the one-call path, `masque()` built the roles table without asking.
Building and reviewing it directly gives full control:

```{r roles}
roles <- propose_roles(df, mode = "collaborate")
knitr::kable(
  roles[, c("col", "role", "action", "kind")],
  caption = paste0(
    "The proposed masking plan for the alpha-design trial ",
    "(collaborate mode)."
  )
)
```

`propose_roles()` fills in a sensible action for each column given the
mode, so the table reviewed here is the plan that will run. `yield` was
proposed as a `covariate`; naming it the trial's `outcome` changes nothing
about how it is masked, but it documents intent and is what a downstream
consumer such as the *conditional* clone (see *Confidentiality and the
threat model*) reads to find the response:

```{r set-role}
roles <- set_role(roles, "yield", role = "outcome")
knitr::kable(
  roles[, c("col", "role", "action")],
  caption = "The plan after naming `yield` as the trial outcome."
)
```

Re-assigning a role re-resolves the default action; passing an explicit
`action` pins the column instead. There is no requirement to name an
outcome: with none marked, every scrambled numeric is re-simulated jointly.

### Editing the plan as code

The printed table -- and the spreadsheet the guided prompt opens when the
custodian chooses `e` -- is an ordinary data frame, so anything the editor
can do, a script can do reproducibly. `set_role()` is vectorised over
columns:

```{r edit-vectorised}
r2 <- set_role(roles, c("row", "col"), action = "drop")
```

Direct edits work too. A direct `role` edit leaves `action` untouched, so
setting the action back to `NA` tells `mask()` to re-resolve the default
for the new role:

```{r edit-direct}
r2$role[r2$col == "rep"] <- "covariate"
r2$action[r2$col == "rep"] <- NA
knitr::kable(
  r2[, c("col", "role", "action")],
  caption = paste0(
    "A hand-edited plan: `row`/`col` dropped, `rep` re-roled to a ",
    "covariate."
  )
)
```

The `kind` column comes from the column's class and cannot be edited. If
it is wrong, convert the column in the data and run `propose_roles()`
again.

`role_options()` lists the pairs the package accepts, here for a factor
column:

```{r role-options}
knitr::kable(
  role_options(kind = "factor"),
  caption = paste0(
    "Every (role, action) pair the validator accepts for a factor ",
    "column."
  )
)
```

Finally, `pii_suspected`. `propose_roles()` sets it from the column
*name* (`email`, `phone`, `owner`, ...), and the leakage audit treats a
flagged column that survives into the synthetic as a HIGH finding. The
scan reads names, not content, so when a harmlessly named column holds
sensitive values, flagging it by hand is the correct fix:

```{r pii-by-hand, eval = FALSE}
roles$pii_suspected[roles$col == "comments"] <- TRUE
```

Passing the edited table back to `mask()` (or to `masque(df, roles =
roles)`) runs the plan:

```{r mask}
m <- mask(df, roles, mode = "collaborate", seed = 1L)
synth <- synthetic(m)
byte_identical_plot <- identical(synth$plot, df$plot)
gen_levels_changed <- !setequal(levels(synth$gen), levels(df$gen))
byte_identical_plot
gen_levels_changed
head(levels(synth$gen))
```

### A plan the package refuses

`plot` is a `design` column. Scrambling it would break the trial's layout,
and `mask()` refuses:

```{r fail-closed, error = TRUE}
bad_roles <- set_role(roles, "plot", role = "design", action = "scramble")
mask(df, bad_roles, mode = "collaborate", seed = 1L)
```

The call stops before any masking, names the column and the combination at
fault, and lists the three actions that do work for a design column
(`keep`, `alias`, `drop`). The same check runs whether the plan came from
`propose_roles()`, a hand edit or the guided spreadsheet.

### More than one table

A confidential dataset often arrives as several related files or a
multi-sheet workbook. `masque()` handles those too -- point it at a
folder, an `.xlsx` file, or a named list of data frames and it masks every
table at once, aliasing any shared key (a site code, a genotype name)
*identically across tables* so a join of the synthetic tables still
resolves.

```{r set}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- masque(set_dir, mode = "collaborate", seed = 1L, ask = FALSE)
ms
```

The genotype column `gen` appears in both tables and is masked to the same
codes in each, so the field and laboratory tables still join. See
*Confidentiality and the threat model* for the set-level controls.

### Multi-environment structure

A multi-environment trial has at least three distinct structural
questions: which rows belong to each environment, whether treatments
connect the environments, and what randomisation structure can be
recovered within each environment. `detect_design()` reports these
separately. Connectivity says whether environments can be compared; it
does not decide whether the trial is multi-environment. A block or field
layout seen in the data suggests the original randomisation but does not
confirm it.

A small synthetic toy with two environments and three genotypes makes the
three questions concrete:

```{r met-scope}
met <- expand.grid(
  env = factor(c("E1", "E2")),
  rep = factor(seq_len(2L)),
  gen = factor(c("G1", "G2", "G3"))
)
met$yield <- seq_len(nrow(met)) + rep(c(0, 2), each = 6L)

ds <- detect_design(met)
ds@scope_label
ds@env_cols
ds@connectivity$status
ds@within_design_label
```

Automatic detection is conservative. A column named `env` or
`environment`, or a common site-year pattern, is taken as the environment
only when no other column explains the structure as well; a site-only
column is taken only when treatments are replicated across sites, so a
nested block is not mistaken for an environment. When you know the
environment columns, name them:

```{r met-explicit}
ds_explicit <- detect_design(met, env = "env")
```

### Figure: the multi-environment coverage plot

`plot.design_summary()` draws a compact environment overview by default,
in base graphics or, with `engine = "ggplot2"`, as a `ggplot2` object you
can save with `ggplot2::ggsave()`; printed, it is the figure below. It
shows whether coverage is even across environments, which the
connectivity flag alone does not.

```{r met-plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.cap = "Genotypes observed per environment (E1, E2) for the toy multi-environment trial; the subtitle carries connectivity status, component count, and the recovered within-environment design."}
plot(ds_explicit, df = met, engine = "ggplot2")
```

```{r met-plot-skip, eval = !requireNamespace("ggplot2", quietly = TRUE), echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the coverage figure is",
    "skipped; `plot(ds_explicit, df = met)` still draws the base-graphics",
    "equivalent.\n")
```

Both bars are at three treatments, since every genotype in this toy trial
appears in both environments. A shorter bar, or a subtitle reporting more
than one component, marks environments that cannot go into one connected
analysis until you have checked whether the missing coverage is real or an
artefact of how the table was put together.

High-confidence environment recommendations feed the masking plan directly.
Local mode keeps environment values byte-identical; collaborate mode
aliases categorical environment labels in place, keeping each row in its
environment and the missing values where they were, and the recipe can
turn the aliases back into the real labels. A numeric
environment such as year remains `keep` and raises a disclosure warning
because its values are still visible:

```{r met-roles}
met_roles <- propose_roles(met, mode = "collaborate")
knitr::kable(
  met_roles[met_roles$col == "env", c("col", "role", "action")],
  caption = "How the masking plan treats the `env` column."
)
```

This safeguard protects the allocation a pipeline reads. It does not imply
that the synthesised outcomes preserve genotype-by-environment effects:
sparse treatment-by-environment cells may fall back to pooled synthesis,
and the clone must not be used as a substitute for the original trial in
scientific inference.

## Read

The audit found nothing to flag on this trial, so the synthetic could be
handed over as it is. A design column left at `keep` comes back unchanged
and an aliased treatment comes back as `trt_001` to `trt_024`; a plan that
asks for something the package cannot do is refused before it runs. Related
tables are masked together and still join.

The agronomist can therefore hand over a clone that keeps the plot layout
and the links between tables, with the genotype labels aliased, and the
statistician can build a full pipeline on it without seeing the real
genotype names. How the finished pipeline runs
on the real data is covered in *Recipe anatomy and the round-trip*.

## Limits

This vignette shows the custodian's side of the handoff, on one small
public trial with a clean, complete design; it does not exercise the
leakage audit's failure modes, the geographic-coordinate controls, or the
conditional clone that preserves a treatment effect rather than only a
marginal distribution, all of which live in *Confidentiality and the
threat model*. A sensitive column with an ordinary name is caught only if
someone flags it by hand. Multi-environment detection can miss a real
environment when the names and replication are ambiguous; prefer an
explicit `env` whenever you know better. Only one numeric column is
re-simulated here, so the vignette does not show what the Gaussian copula
keeps of the relationships between columns; that is also in
*Confidentiality and the threat model*.

## What to read next

*Confidentiality and the threat model* sets out exactly what a synthetic
clone protects and what it does not, the depth controls beyond role and
action, the leakage audit, and the conditional clone that preserves a
treatment-to-outcome relationship. *Recipe anatomy and the round-trip*
is the analyst's side of this handoff: what the recipe carries, and how a
pipeline built on the synthetic re-targets to the original.

## Reproduce

`set.seed(1)` is set once for the document, and every `masque()` or
`mask()` call passes `seed = 1L`, so each is reproducible on its own.
Package versions follow.

```{r session-info}
sessionInfo()
```
