---
title: "Recipe anatomy and the round-trip"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recipe anatomy and the round-trip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS arm64, R 4.5.2, one
     core. No sampler runs in this vignette, so the time is dominated by
     knitr overhead and one small copula fit. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
options(width = 100)
library(masque)
set.seed(1)

f <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
roles <- propose_roles(df, mode = "collaborate")
roles <- set_role(roles, "yield", role = "outcome")
```

## Why

An analyst has the synthetic table and their own code; the custodian keeps
the original and the recipe. This vignette covers what the recipe carries,
how the analyst's finished pipeline is run on the original, and what
happens when the original holds a value the recipe has never seen.

## What

The `masque_recipe` object returned by `mask()` is the only artefact
that *must* stay confidential alongside the original -- it is an S7
object, accessed through `recipe()`, never touched directly. Two verbs
carry the round-trip: `apply_recipe()` translates a data frame from the
original namespace into the synthetic namespace (the direction a
custodian uses to re-target an analyst's finished pipeline), and
`unmask()` translates the other way (the direction that recovers
original-namespace labels from a synthetic-namespace result).
`save_recipe()`/`read_recipe()` persist a recipe to a single `.rds`
file, and `reveal_maps()` is the one explicit, warning-gated call that
shows the level maps a redacted `print()` withholds.

## Do

### What a recipe holds

```{r accessors}
m <- mask(df, roles, mode = "collaborate", seed = 1L)
rec <- recipe(m)
class(rec)
```

A recipe carries only what the round-trip needs:

- `masque_version`, `created_at`, `mode`, `seed` -- provenance.
- `roles` -- the two-axis role and action table.
- `level_maps` -- the per-column original-to-alias maps. The sensitive
  part.
- `column_name_map` -- the column-name aliases, when `alias_names` was
  used (otherwise `NULL`).
- `cleaning` -- the hygiene record (name legalisation, whitespace trims)
  so `apply_recipe()` re-applies the identical tidying.
- `storage_classes`, `factor_meta` -- the original column classes and
  factor metadata, for faithful reconstruction.
- `coords` -- one entry per coordinate pair declared to `mask()`'s
  `coords` argument: the jitter method and radius, the site grouping it
  was masked under, and how many sites the grouping produced. Empty for a
  recipe written before 0.10.0, which had no such record.
- `allow_unmasked_coords` -- `TRUE` only when the caller deliberately
  wrote a real coordinate through unmasked. `FALSE`, and absent, on a
  recipe written before 0.11.0.
- `integrity_fp` -- a SHA-256 of `is.na(original)`: an integrity
  fingerprint, not a privacy guarantee.

It deliberately does not hold the copula covariance, the raw observed
values, or any file paths or usernames.

A recipe built with `coords` carries its own coordinate account, so an
audit of the recipe does not need the original data back:

```{r coords-account, eval = requireNamespace("maps", quietly = TRUE)}
df_geo <- df
df_geo$lat <- -34.9 + stats::runif(nrow(df_geo), -0.05, 0.05)
df_geo$lon <- 138.6 + stats::runif(nrow(df_geo), -0.05, 0.05)
roles_geo <- propose_roles(df_geo, mode = "collaborate")
roles_geo <- set_role(roles_geo, "yield", role = "outcome")
roles_geo <- set_role(roles_geo, c("lat", "lon"), action = "keep")

m_geo <- mask(df_geo, roles_geo, mode = "collaborate", seed = 1L,
              coords = list(list(lat = "lat", lon = "lon")))
n_sites_recorded <- recipe(m_geo)@coords[[1]]$n_sites
recipe(m_geo)@coords[[1]][c("method", "min_km", "max_km", "n_sites")]
recipe(m_geo)@allow_unmasked_coords
```

```{r coords-account-skip, eval = !requireNamespace("maps", quietly = TRUE), echo = FALSE, results = "asis"}
cat("maps is not installed on this build, so the coordinate-account demo",
    "is skipped.\n")
n_sites_recorded <- NA_integer_
```

Each plot here has its own coordinate, so each plot is a site and
`n_sites` equals the number of plots. Rows that share a coordinate would
share one displacement (see `jitter_coordinates()`).

### Printing is redacted

The print method shows the role and action table with a marker for which
columns hold a level map (`*` mapped, `=` not), but never the vocabularies
themselves:

```{r print-redacted}
rec
```

The custodian -- and only the custodian -- can reveal the maps with an
explicit, warning-gated call:

```{r reveal, results = "hide"}
reveal_maps(rec)
```

### The round-trip

The recipe is a bidirectional translator. `apply_recipe()` carries the
original *into* the synthetic namespace. `unmask()` carries results *back*.
The usual pattern is to fit on the synthetic, then score the original:

```{r roundtrip-pipeline}
# Analyst trains on the synthetic clone.
fit <- lm(yield ~ gen, data = synthetic(m))

# Custodian re-targets the same pipeline onto the real data.
orig_in_synth_space <- apply_recipe(df, rec)
preds <- predict(fit, newdata = orig_in_synth_space)
preds_match_nrow <- length(preds) == nrow(df)
preds_match_nrow
```

Recovering original-namespace labels is symmetric:

```{r roundtrip-labels}
fwd <- apply_recipe(df, rec)
back <- unmask(fwd, rec)
labels_recovered <- identical(as.character(back$gen), as.character(df$gen))
labels_recovered
```

Columns whose action is `keep` pass through all three verbs unchanged.
Numeric columns are passed through `apply_recipe()` and `unmask()` without
an inverse map -- the synthetic and original namespaces coincide for
them -- so `unmask()` on a numeric prediction vector is a safe no-op.

### Fail-closed translation, and reading the refusal

If the data drifts, for example a new treatment level the recipe has never
seen, `apply_recipe()` stops with an error:

```{r fail-closed, error = TRUE}
drifted <- df
levels(drifted$gen) <- c(levels(df$gen), "BRAND_NEW")
drifted$gen[1] <- "BRAND_NEW"
apply_recipe(drifted, rec)
```

The message names the column (`gen`) and the value (`"BRAND_NEW"`). Rebuild
the recipe from data that contains the new value, or remove the drifted
rows before re-targeting.

### Saving the recipe

`save_recipe()` writes a single small `.rds`, to be stored next to the
original at the same security class. `read_recipe()`
validates the file and notes a version mismatch without raising an error.

```{r save}
tmp <- tempfile(fileext = ".rds")
save_recipe(rec, tmp)
rec2 <- read_recipe(tmp)
fingerprint_survives_roundtrip <- identical(rec@integrity_fp, rec2@integrity_fp)
fingerprint_survives_roundtrip
```

### Multi-table bundles

A `mask_set()` result carries a recipe *bundle* -- one recipe per table
plus the shared cross-table link maps. The same `apply_recipe()` and
`unmask()` verbs dispatch on it, operating table by table over a named
list:

```{r set-bundle}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- mask_set(set_dir, mode = "collaborate", seed = 1L, quiet = TRUE)
originals <- read_set(set_dir)
fwd_set <- suppressWarnings(apply_recipe(originals, recipe(ms)))
names(fwd_set)
```

Because the shared keys were aliased consistently at masking time, the
re-targeted tables join on the same keys the synthetic tables do.

### Figure: the clone the analyst actually develops against

The checks above are about labels. The figure shows the numeric column the
analyst works with: the original yield against the synthetic yield.

```{r fig-overlay, eval = requireNamespace("ggplot2", quietly = TRUE), fig.cap = "Density of yield, original trial against the synthetic clone the analyst develops against. The two distributions overlap closely because the default numeric synthesis preserves each column's marginal distribution."}
synth_tbl <- synthetic(m)
overlay_df <- rbind(
  data.frame(value = df$yield, table = "original"),
  data.frame(value = synth_tbl$yield, table = "synthetic")
)
ggplot2::ggplot(overlay_df, ggplot2::aes(x = value, fill = table)) +
  ggplot2::geom_density(alpha = 0.5) +
  ggplot2::scale_fill_viridis_d(name = "table") +
  ggplot2::labs(
    x = "yield (t/ha)", y = "density",
    title = "Original versus synthetic marginal distribution of yield"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-overlay-skip, eval = !requireNamespace("ggplot2", quietly = TRUE), echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the marginal-overlay",
    "figure is skipped.\n")
```

## Read

All three round-trip checks pass: one prediction per original row, the
original genotype labels recovered, and the same integrity fingerprint
after the recipe is saved and read back. A value the recipe has never seen
stops the translation where the data drifted.

The two yield densities overlap closely, so the marginal distribution is
kept. That says nothing about whether a treatment effect or a non-linear
relationship between columns is kept; see *Confidentiality and the threat
model*.

With the recipe, the custodian can run the analyst's finished pipeline on
the real data without changing it.

## Limits

This vignette uses one recipe built from one clean public trial, with one
planted drift. A table with inconsistent historical vocabularies will be
refused more often. The round-trip verbs invert only what the recipe
recorded: numeric columns pass through unchanged, so `unmask()` cannot undo
a numeric transformation a pipeline made, and a recipe cannot repair a join
that was broken before masking. This vignette does not exercise the leakage audit,
the conditional clone, or the geographic-coordinate masking controls in
depth; those, and what the recipe alone cannot protect against a
determined reader of the synthetic, are *Confidentiality and the threat
model*'s subject.

## What to read next

*Getting started with masque* is the custodian's side of this same
handoff: building and reviewing the masking plan that produces the recipe
this vignette reads. *Confidentiality and the threat model* sets out what
the recipe and the synthetic together do and do not protect, including the
leakage audit and the conditional clone that preserves a treatment effect
this vignette's round-trip checks do not touch.

## Reproduce

`set.seed(1)` is set once for the document, and every `mask()` or
`mask_set()` call passes `seed = 1L`, so each is reproducible on its own.
Package versions follow.

```{r session-info}
sessionInfo()
```
