---
title: "Recipe anatomy and the round-trip"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recipe anatomy and the round-trip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: ~2 s under rmarkdown::render(); macOS arm64, R 4.5.2, one
     core. No sampler runs in this vignette, so the time is dominated by
     knitr overhead and one small copula fit. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 150
)
options(width = 100)
library(masque)
set.seed(1)

f <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
roles <- propose_roles(df, mode = "collaborate")
roles <- set_role(roles, "yield", role = "outcome")
```

## Why

An analyst has only the synthetic table and their own code -- the
custodian kept the original and the recipe. Their question is the one
every downstream user of a `masque` clone eventually asks: *what do I
actually get back from the custodian, and what happens the day the real
data does not quite match what I built against?* This vignette is written
for that analyst's side of the workflow: what the recipe carries, how the
round-trip works, and what it does when the two sides of the trust
boundary disagree.

## What

The `masque_recipe` object returned by `mask()` is the only artefact
that *must* stay confidential alongside the original -- it is an S7
object, accessed through `recipe()`, never touched directly. Two verbs
carry the round-trip: `apply_recipe()` translates a data frame from the
original namespace into the synthetic namespace (the direction a
custodian uses to re-target an analyst's finished pipeline), and
`unmask()` translates the other way (the direction that recovers
original-namespace labels from a synthetic-namespace result).
`save_recipe()`/`read_recipe()` persist a recipe to a single `.rds`
file, and `reveal_maps()` is the one explicit, warning-gated call that
shows the level maps a redacted `print()` withholds.

## Do

### What a recipe holds

```{r accessors}
m <- mask(df, roles, mode = "collaborate", seed = 1L)
rec <- recipe(m)
class(rec)
```

A recipe is runtime-minimal by default. It carries:

- `masque_version`, `created_at`, `mode`, `seed` -- provenance.
- `roles` -- the two-axis role and action table.
- `level_maps` -- the per-column original-to-alias maps. The sensitive
  part.
- `column_name_map` -- the column-name aliases, when `alias_names` was
  used (otherwise `NULL`).
- `cleaning` -- the hygiene record (name legalisation, whitespace trims)
  so `apply_recipe()` re-applies the identical tidying.
- `storage_classes`, `factor_meta` -- the original column classes and
  factor metadata, for faithful reconstruction.
- `coords` -- one entry per coordinate pair declared to `mask()`'s
  `coords` argument: the jitter method and radius, the site grouping it
  was masked under, and how many sites the grouping produced. Empty for a
  recipe written before 0.10.0, which had no such record.
- `allow_unmasked_coords` -- `TRUE` only when the caller deliberately
  wrote a real coordinate through unmasked. `FALSE`, and absent, on a
  recipe written before 0.11.0.
- `integrity_fp` -- a SHA-256 of `is.na(original)`: an integrity
  fingerprint, not a privacy guarantee.

It deliberately does not hold the copula covariance, the raw observed
values, or any file paths or usernames.

A recipe built with `coords` carries its own coordinate account, so an
audit of the recipe does not need the original data back:

```{r coords-account, eval = requireNamespace("maps", quietly = TRUE)}
df_geo <- df
df_geo$lat <- -34.9 + stats::runif(nrow(df_geo), -0.05, 0.05)
df_geo$lon <- 138.6 + stats::runif(nrow(df_geo), -0.05, 0.05)
roles_geo <- propose_roles(df_geo, mode = "collaborate")
roles_geo <- set_role(roles_geo, "yield", role = "outcome")
roles_geo <- set_role(roles_geo, c("lat", "lon"), action = "keep")

m_geo <- mask(df_geo, roles_geo, mode = "collaborate", seed = 1L,
              coords = list(list(lat = "lat", lon = "lon")))
n_sites_recorded <- recipe(m_geo)@coords[[1]]$n_sites
recipe(m_geo)@coords[[1]][c("method", "min_km", "max_km", "n_sites")]
recipe(m_geo)@allow_unmasked_coords
```

```{r coords-account-skip, eval = !requireNamespace("maps", quietly = TRUE), echo = FALSE, results = "asis"}
cat("maps is not installed on this build, so the coordinate-account demo",
    "is skipped.\n")
n_sites_recorded <- NA_integer_
```

Every row here shared one true coordinate, so `n_sites` equals the row
count -- the recipe records that the whole table was jittered as one
group, not each row independently, which is the confidentiality property
`jitter_coordinates(by = )` is for.

### Printing is redacted

The print method shows the role and action table with a marker for which
columns hold a level map (`*` mapped, `=` not), but never the vocabularies
themselves:

```{r print-redacted}
rec
```

The custodian -- and only the custodian -- can reveal the maps with an
explicit, warning-gated call:

```{r reveal, results = "hide"}
reveal_maps(rec)
```

### The round-trip

The recipe is a bidirectional translator. `apply_recipe()` carries the
original *into* the synthetic namespace. `unmask()` carries results *back*.
The usual pattern is to fit on the synthetic, then score the original:

```{r roundtrip-pipeline}
# Analyst trains on the synthetic clone.
fit <- lm(yield ~ gen, data = synthetic(m))

# Custodian re-targets the same pipeline onto the real data.
orig_in_synth_space <- apply_recipe(df, rec)
preds <- predict(fit, newdata = orig_in_synth_space)
preds_match_nrow <- length(preds) == nrow(df)
preds_match_nrow
```

Recovering original-namespace labels is symmetric:

```{r roundtrip-labels}
fwd <- apply_recipe(df, rec)
back <- unmask(fwd, rec)
labels_recovered <- identical(as.character(back$gen), as.character(df$gen))
labels_recovered
```

Columns whose action is `keep` pass through all three verbs unchanged.
Numeric columns are passed through `apply_recipe()` and `unmask()` without
an inverse map -- the synthetic and original namespaces coincide for
them -- so `unmask()` on a numeric prediction vector is a safe no-op.

### Fail-closed translation, and reading the refusal

If the data drifts -- a new treatment level the recipe has never seen --
`apply_recipe()` errors rather than silently coercing the unknown value to
`NA`:

```{r fail-closed, error = TRUE}
drifted <- df
levels(drifted$gen) <- c(levels(df$gen), "BRAND_NEW")
drifted$gen[1] <- "BRAND_NEW"
apply_recipe(drifted, rec)
```

Reading the message: it names the exact column (`gen`) and the exact
unrecognised value (`"BRAND_NEW"`), states plainly that unknown values are
not silently coerced to `NA`, and gives the remedy -- rebuild the recipe
from a dataset that contains the new value, or strip the drifted rows
before re-targeting. This is deliberate: a model matrix built from a
silently-introduced `NA` would fail somewhere downstream with a far less
legible error, or worse, would not fail at all and would simply drop the
drifted rows. Schema drift is exactly the kind of change that should stop
a pipeline at the point it happens, not several steps later.

### Saving the recipe

`save_recipe()` writes a single runtime-minimal `.rds` -- small, and safe
to store next to the original at the same security class. `read_recipe()`
validates the file and notes a version mismatch without raising an error.

```{r save}
tmp <- tempfile(fileext = ".rds")
save_recipe(rec, tmp)
rec2 <- read_recipe(tmp)
fingerprint_survives_roundtrip <- identical(rec@integrity_fp, rec2@integrity_fp)
fingerprint_survives_roundtrip
```

### Multi-table bundles

A `mask_set()` result carries a recipe *bundle* -- one recipe per table
plus the shared cross-table link maps. The same `apply_recipe()` and
`unmask()` verbs dispatch on it, operating table by table over a named
list:

```{r set-bundle}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- mask_set(set_dir, mode = "collaborate", seed = 1L, quiet = TRUE)
originals <- read_set(set_dir)
fwd_set <- suppressWarnings(apply_recipe(originals, recipe(ms)))
names(fwd_set)
```

Because the shared keys were aliased consistently at masking time, the
re-targeted tables join on the same keys the synthetic tables do.

### Figure: the clone the analyst actually develops against

Everything above works with the *labels*. The figure below shows what the
analyst's numeric column actually looks like -- the original yield
against the synthetic yield the analyst's pipeline is fitted to -- which
the round-trip checks above, all of them logical values, cannot show.

```{r fig-overlay, eval = requireNamespace("ggplot2", quietly = TRUE), fig.cap = "Density of yield, original trial against the synthetic clone the analyst develops against. The two distributions overlap closely because the default numeric synthesis preserves each column's marginal distribution."}
synth_tbl <- synthetic(m)
overlay_df <- rbind(
  data.frame(value = df$yield, table = "original"),
  data.frame(value = synth_tbl$yield, table = "synthetic")
)
ggplot2::ggplot(overlay_df, ggplot2::aes(x = value, fill = table)) +
  ggplot2::geom_density(alpha = 0.5) +
  ggplot2::scale_fill_viridis_d(name = "table") +
  ggplot2::labs(
    x = "yield (t/ha)", y = "density",
    title = "Original versus synthetic marginal distribution of yield"
  ) +
  ggplot2::theme_minimal()
```

```{r fig-overlay-skip, eval = !requireNamespace("ggplot2", quietly = TRUE), echo = FALSE, results = "asis"}
cat("ggplot2 is not installed on this build, so the marginal-overlay",
    "figure is skipped.\n")
```

## Read

The round-trip holds on every check run above: the re-targeted prediction
vector has one entry per original row
(`r isTRUE(preds_match_nrow)`), the recovered genotype labels are
identical to the original after a forward-then-back trip through the
recipe (`r isTRUE(labels_recovered)`), and a recipe written to disk and
read back carries an unchanged integrity fingerprint
(`r isTRUE(fingerprint_survives_roundtrip)`). The coordinate account
records `r n_sites_recorded` sites for a fixture where every plot shared
one true coordinate, which is the row count of the trial -- confirming the
whole table was jittered as one group rather than plot by plot. The
fail-closed section shows the other side of the same reliability property:
a value the recipe has never seen is refused, not guessed at, so a
pipeline re-targeted through `apply_recipe()` fails at the point the data
actually drifted rather than downstream with a harder-to-diagnose error.

The figure shows why the round-trip's logical checks are not the whole
story: the two densities overlap closely, which is what "the marginal
distribution is preserved" looks like on real data, but overlap in the
marginal says nothing about whether a treatment effect or a non-linear
covariate relationship survived -- for that, see *Confidentiality and the
threat model*.

Answering the opening question: the recipe an analyst is handed back after
their pipeline is finished lets a custodian re-target that exact pipeline,
unmodified, onto the real data, and the moment the real data carries
something the recipe was never built from, the translation stops rather
than guessing.

## Limits

Everything in this vignette exercises one recipe built from one clean
public trial with no missing structure and no schema drift beyond the one
planted for the fail-closed demonstration; a recipe built from a table
with genuinely inconsistent historical vocabularies will refuse more
often, by design, and the remedy each time is the same -- rebuild from
data that contains the missing values, or strip the drifted rows. The
round-trip verbs invert exactly what the recipe recorded: a numeric column
has no inverse map and passes through unchanged, so `unmask()` cannot
recover anything about a numeric transformation a pipeline applied on its
own, and a recipe cannot repair a join that was broken outside `masque`
before masking ever ran. This vignette does not exercise the leakage audit,
the conditional clone, or the geographic-coordinate masking controls in
depth; those, and what the recipe alone cannot protect against a
determined reader of the synthetic, are *Confidentiality and the threat
model*'s subject.

## What to read next

*Getting started with masque* is the custodian's side of this same
handoff: building and reviewing the masking plan that produces the recipe
this vignette reads. *Confidentiality and the threat model* sets out what
the recipe and the synthetic together do and do not protect, including the
leakage audit and the conditional clone that preserves a treatment effect
this vignette's round-trip checks do not touch.

## Reproduce

`set.seed(1)` is set once for the whole document, and every `mask()`/
`mask_set()` call also passes `seed = 1L` explicitly, so each is
independently reproducible regardless of what ran before it in this
vignette. Package versions follow.

```{r session-info}
sessionInfo()
```
