---
title: "Recipe anatomy and the round-trip"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recipe anatomy and the round-trip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
library(masque)
set.seed(1)

f <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
roles <- propose_roles(df, mode = "collaborate")
roles <- set_role(roles, "yield", role = "outcome")
```

## Why the recipe matters

This vignette is written for the analyst's side of the workflow. The
custodian masks the data and keeps the recipe. You receive only the
synthetic. But the recipe is what makes a pipeline you write against the
synthetic re-target to the real data, so it is worth understanding what
it carries and how the round-trip works.

The `masque_recipe` is the only artefact that *must* stay confidential
alongside the original. It is an S7 object, but you never touch the class
directly -- two accessors hide it:

```{r accessors}
m <- mask(df, roles, mode = "collaborate", seed = 1)
rec <- recipe(m)
class(rec)
```

## What a recipe holds

A recipe is runtime-minimal by default:

- `masque_version`, `created_at`, `mode`, `seed` -- provenance.
- `roles` -- the two-axis role and action table.
- `level_maps` -- the per-column original-to-alias maps. The sensitive
  part.
- `column_name_map` -- the column-name aliases, when `alias_names` was
  used (otherwise `NULL`).
- `cleaning` -- the hygiene record (name legalisation, whitespace trims)
  so `apply_recipe()` re-applies the identical tidying.
- `storage_classes`, `factor_meta` -- the original column classes and
  factor metadata, for faithful reconstruction.
- `coords` -- one entry per coordinate pair declared to `mask()`'s
  `coords` argument: the jitter method and radius, the site grouping it
  was masked under, and how many sites the grouping produced. Empty for
  a recipe written before 0.10.0, which had no such record.
- `allow_unmasked_coords` -- `TRUE` only when the caller deliberately
  wrote a real coordinate through unmasked. `FALSE`, and absent, on a
  recipe written before 0.11.0.
- `integrity_fp` -- a SHA-256 of `is.na(original)`: an integrity
  fingerprint, not a privacy guarantee.

It deliberately does not hold the copula covariance, the raw observed
values, or any file paths or usernames.

A recipe built with `coords` carries its own coordinate account, so an
audit of the recipe does not need the original data back:

```{r coords-account, eval = requireNamespace("maps", quietly = TRUE)}
df_geo <- df
df_geo$lat <- -34.9 + stats::runif(nrow(df_geo), -0.05, 0.05)
df_geo$lon <- 138.6 + stats::runif(nrow(df_geo), -0.05, 0.05)
roles_geo <- propose_roles(df_geo, mode = "collaborate")
roles_geo <- set_role(roles_geo, "yield", role = "outcome")
roles_geo <- set_role(roles_geo, c("lat", "lon"), action = "keep")

m_geo <- mask(df_geo, roles_geo, mode = "collaborate", seed = 1,
              coords = list(list(lat = "lat", lon = "lon")))
recipe(m_geo)@coords[[1]][c("method", "min_km", "max_km", "n_sites")]
recipe(m_geo)@allow_unmasked_coords
```

Every row here shared one true coordinate, so `n_sites` above equals the
row count -- the recipe records that the whole table was jittered as one
group, not each row independently, which is the confidentiality property
`jitter_coordinates(by = )` is for.

## Printing is redacted

The print method shows the role and action table with a marker for which
columns hold a level map (`*` mapped, `=` not), but never the
vocabularies themselves:

```{r print-redacted}
rec
```

The custodian -- and only the custodian -- can reveal the maps with an
explicit, warning-gated call:

```{r reveal, results = "hide"}
reveal_maps(rec)
```

## The round-trip

The recipe is a bidirectional translator. `apply_recipe()` carries the
original *into* the synthetic namespace. `unmask()` carries results
*back*. The usual pattern is to fit on the synthetic, then score the
original:

```{r roundtrip-pipeline}
# Analyst trains on the synthetic clone.
fit <- lm(yield ~ gen, data = synthetic(m))

# Custodian re-targets the same pipeline onto the real data.
orig_in_synth_space <- apply_recipe(df, rec)
preds <- predict(fit, newdata = orig_in_synth_space)
length(preds) == nrow(df)
```

Recovering original-namespace labels is symmetric:

```{r roundtrip-labels}
fwd <- apply_recipe(df, rec)
back <- unmask(fwd, rec)
identical(as.character(back$gen), as.character(df$gen))
```

Columns whose action is `keep` pass through all three verbs unchanged.
Numeric columns are passed through `apply_recipe()` and `unmask()`
without an inverse map -- the synthetic and original namespaces coincide
for them -- so `unmask()` on a numeric prediction vector is a safe no-op.

## Fail-closed translation

If the data drifts -- a new treatment level the recipe has never seen --
`apply_recipe()` errors rather than silently coercing the unknown value
to `NA`. Schema drift should fail loudly, not poison a model matrix.

```{r fail-closed, error = TRUE}
drifted <- df
levels(drifted$gen) <- c(levels(df$gen), "BRAND_NEW")
drifted$gen[1] <- "BRAND_NEW"
apply_recipe(drifted, rec)
```

## Saving the recipe

`save_recipe()` writes a single runtime-minimal `.rds` -- small, and safe
to store next to the original at the same security class. `read_recipe()`
validates the file and notes a version mismatch without raising an error.

```{r save}
tmp <- tempfile(fileext = ".rds")
save_recipe(rec, tmp)
rec2 <- read_recipe(tmp)
identical(rec@integrity_fp, rec2@integrity_fp)
```

## Multi-table bundles

A `mask_set()` result carries a recipe *bundle* -- one recipe per table
plus the shared cross-table link maps. The same `apply_recipe()` and
`unmask()` verbs dispatch on it, operating table by table over a named
list:

```{r set-bundle}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- mask_set(set_dir, mode = "collaborate", seed = 1, quiet = TRUE)
originals <- read_set(set_dir)
fwd_set <- apply_recipe(originals, recipe(ms))
names(fwd_set)
```

Because the shared keys were aliased consistently at masking time, the
re-targeted tables join on the same keys the synthetic tables do.
