---
title: "Recipe anatomy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recipe anatomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
library(masque)

f  <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
roles <- propose_roles(df)
roles$role[roles$col == "yield"] <- "outcome"
```

# What is a recipe?

The `masque_recipe` is the only artefact that *must* remain confidential
alongside the original data. It is what makes the round-trip work:
without the recipe you cannot translate a pipeline trained on the
synthetic back onto the original.

A recipe is an S7 object, but users do not need to know that. The two
exported accessors hide the class details:

```{r accessors}
m   <- mask(df, roles, mode = "collaborate", seed = 1)
rec <- recipe(m)
class(rec)
```

## Anatomy

A recipe holds **runtime-minimal** state by default:

- `masque_version` — the package version that built the recipe.
- `created_at` — wall-clock timestamp at construction.
- `mode` — `"local"` or `"collaborate"`.
- `seed` — the seed passed to `mask()`, or `NULL` if not given.
- `roles` — the per-column role tibble.
- `column_name_map` — original-to-synthetic column-name map (NULL in
  v0.2; reserved for v0.3+).
- `level_maps` — per-column factor / character maps. The sensitive bit.
- `storage_classes` — per-column R class of the original.
- `factor_meta` — per-factor levels and `ordered` status.
- `warnings` — text of any warnings raised at construction.
- `integrity_fp` — SHA-256 of `is.na(original)`. An **integrity
  fingerprint**, not a privacy guarantee.

What it deliberately does **not** hold:

- Simulator state (the copula covariance matrix, the raw observed
  margins). This is a v0.3+ extension behind
  `save_recipe(..., include_simulator = TRUE)`; currently a no-op.
- Raw observed values.
- Source file paths, machine usernames, or absolute paths.

## `print(recipe)` is redacted

The default print method shows the per-column role table and a marker
indicating whether a level map exists for each column (`*` = mapped,
`=` = no map), but **never** the actual level vocabularies.

```{r print-redacted}
rec
```

If you need to inspect the maps — typically the data owner reviewing
the recipe before saving — call `reveal_maps()` explicitly:

```{r reveal, results = "hide"}
reveal_maps(rec)
```

`reveal_maps()` prints a warning banner ("Revealing sensitive level
maps. Proceed at your discretion.") and then dumps every map and the
seed value. Save its output sparingly.

## Saving and loading

`save_recipe()` writes a single `.rds` file. The default is
runtime-minimal — small, safe to store next to the original data with
the same security class.

```{r save}
tmp <- tempfile(fileext = ".rds")
save_recipe(rec, tmp)
file.info(tmp)$size
```

`read_recipe()` validates the file and informs (does not error) when
the recorded `masque_version` differs from the currently installed
package version.

```{r read}
rec2 <- read_recipe(tmp)
identical(rec@integrity_fp, rec2@integrity_fp)
```

## The integrity fingerprint

`integrity_fp` is `digest::digest(is.na(original), algo = "sha256")`.
It lets a downstream consumer check that a recipe corresponds to the
expected missingness pattern, without exposing any other information
about the original data.

```{r integrity}
digest::digest(is.na(df), algo = "sha256") == rec@integrity_fp
```

It is **not** a privacy mechanism. The hash tells you whether two
data frames share the same NA mask; it does not hide the underlying
mask or its risks.

## Round-trip the maps directly

The recipe is the bidirectional translator. `apply_recipe()` and
`unmask()` both operate on it:

```{r roundtrip}
fwd  <- apply_recipe(df, rec)
back <- unmask(fwd, rec)
identical(as.character(back$gen), as.character(df$gen))
```

## Future: `include_simulator = TRUE`

`save_recipe(rec, path, include_simulator = TRUE)` is accepted today
but is currently a no-op (no simulator state is stored on the recipe
in v0.2). v0.3+ will use this flag to persist enough state that
`draw_new_synthetic(rec, n)` can produce fresh synthetic samples
without access to the original. See `vignette("roadmap")` for the
deferred items.
