---
title: "Recipe anatomy and the round-trip"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recipe anatomy and the round-trip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 100)
library(masque)
set.seed(1)

f <- system.file("extdata", "john_alpha.csv", package = "masque")
df <- read.csv(f, stringsAsFactors = TRUE)
roles <- propose_roles(df, mode = "collaborate")
roles <- set_role(roles, "yield", role = "outcome")
```

## Why the recipe matters

This vignette is written for the analyst's side of the workflow. The
custodian masks the data and keeps the recipe; you receive only the
synthetic. But the recipe is what makes a pipeline you write against the
synthetic re-target to the real data, so it is worth understanding what
it carries and how the round-trip works.

The `masque_recipe` is the only artefact that *must* stay confidential
alongside the original. It is an S7 object, but you never touch the class
directly -- two accessors hide it:

```{r accessors}
m <- mask(df, roles, mode = "collaborate", seed = 1)
rec <- recipe(m)
class(rec)
```

## What a recipe holds

A recipe is runtime-minimal by default:

- `masque_version`, `created_at`, `mode`, `seed` -- provenance.
- `roles` -- the two-axis role and action table.
- `level_maps` -- the per-column original-to-alias maps. The sensitive
  part.
- `column_name_map` -- the column-name aliases, when `alias_names` was
  used (otherwise `NULL`).
- `cleaning` -- the hygiene record (name legalisation, whitespace trims)
  so `apply_recipe()` re-applies the identical tidying.
- `storage_classes`, `factor_meta` -- the original column classes and
  factor metadata, for faithful reconstruction.
- `integrity_fp` -- a SHA-256 of `is.na(original)`: an integrity
  fingerprint, not a privacy guarantee.

It deliberately does not hold the copula covariance, the raw observed
values, or any file paths or usernames.

## Printing is redacted

The print method shows the role and action table with a marker for which
columns hold a level map (`*` mapped, `=` not), but never the
vocabularies themselves:

```{r print-redacted}
rec
```

The custodian -- and only the custodian -- can reveal the maps with an
explicit, warning-gated call:

```{r reveal, results = "hide"}
reveal_maps(rec)
```

## The round-trip

The recipe is a bidirectional translator. `apply_recipe()` carries the
original *into* the synthetic namespace; `unmask()` carries results
*back*. The usual pattern is to fit on the synthetic, then score the
original:

```{r roundtrip-pipeline}
# Analyst trains on the synthetic clone.
fit <- lm(yield ~ gen, data = synthetic(m))

# Custodian re-targets the same pipeline onto the real data.
orig_in_synth_space <- apply_recipe(df, rec)
preds <- predict(fit, newdata = orig_in_synth_space)
length(preds) == nrow(df)
```

Recovering original-namespace labels is symmetric:

```{r roundtrip-labels}
fwd <- apply_recipe(df, rec)
back <- unmask(fwd, rec)
identical(as.character(back$gen), as.character(df$gen))
```

Columns whose action is `keep` pass through all three verbs unchanged.
Numeric columns are passed through `apply_recipe()` and `unmask()`
without an inverse map -- the synthetic and original namespaces coincide
for them -- so `unmask()` on a numeric prediction vector is a safe no-op.

## Fail-closed translation

If the data drifts -- a new treatment level the recipe has never seen --
`apply_recipe()` errors rather than silently coercing the unknown value
to `NA`. Schema drift should fail loudly, not poison a model matrix.

```{r fail-closed, error = TRUE}
drifted <- df
levels(drifted$gen) <- c(levels(df$gen), "BRAND_NEW")
drifted$gen[1] <- "BRAND_NEW"
apply_recipe(drifted, rec)
```

## Saving the recipe

`save_recipe()` writes a single runtime-minimal `.rds` -- small, and safe
to store next to the original at the same security class. `read_recipe()`
validates the file and notes a version mismatch without raising an error.

```{r save}
tmp <- tempfile(fileext = ".rds")
save_recipe(rec, tmp)
rec2 <- read_recipe(tmp)
identical(rec@integrity_fp, rec2@integrity_fp)
```

## Multi-table bundles

A `mask_set()` result carries a recipe *bundle* -- one recipe per table
plus the shared cross-table link maps. The same `apply_recipe()` and
`unmask()` verbs dispatch on it, operating table by table over a named
list:

```{r set-bundle}
set_dir <- system.file("extdata", "met_set", package = "masque")
ms <- mask_set(set_dir, mode = "collaborate", seed = 1, quiet = TRUE)
originals <- read_set(set_dir)
fwd_set <- apply_recipe(originals, recipe(ms))
names(fwd_set)
```

Because the shared keys were aliased consistently at masking time, the
re-targeted tables join on the same keys the synthetic tables do.
